### 2015-05-25 paper/paper.tex
paper and stuff

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\DeclareMathOperator*{\Exp}{\mathbb{E}}
\DeclareMathOperator*{\Prob}{\mathbf{P}}

\title{Qualitative Probabilistic Programming}

\author{
Jessica Taylor \\
Department of Computer Science\\
Stanford University \\
Stanford, CA 94305
\texttt{jessica.liu.taylor@gmail.com}
\And
Noah Goodman \\
Department of ?? \\
Stanford University \\
Stanford, CA 94305
\texttt{...}
\And
}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

  \maketitle

  \begin{abstract}
    In probabilistic programs, sometimes it is difficult to specify the correct
    parameterized family of distributions.  We explore an extension to
    probabilistic programming languages that allows programmers to mark some
    distributions as unspecified.  Then, we can fill in the distribution with
    some family and infer parameters.
  \end{abstract}

  \section{INTRODUCTION}

  By separating model specification and inference, probabilistic programming 
  has made it easier for non-experts to implement and use probabilistic models.
  Practitioners frequently have strong intuitions about the {\em structure}
  of their domain knowledge, such as which latent variables exist and what
  their causal relations are, and probabilistic programming allows them to encode
  this knowledge. However, it also requires them to specify the specific parametric
  shape  and parameterization of any distributions used, and intuitions tend to
  be much less precise there.
  We present Quipp, a system that does {\em not} require such specification;
  instead, random variables and random functions can be left undefined
  and will automatically be filled in under maximum entropy assumptions
  based on their types and available datasets.

  Our formalism can concisely express a wide variety of models that machine
  learning practitioners care about, and we provide an expectation
  maximization algorithm that can learn the parameters for many of
  these models with reasonable efficiency. This system makes it easy
  for non-experts to encode their beliefs about the data and to get
  predictions based on as few additional assumptions as possible.

  % This feature has multiple advantages.  First, it is easier to write
  % models without knowing about the class of models being used.  This should
  % make probabliistic programming more accessible to non-experts.  Secondly,
  % parameter inference is more efficient if specialized algorithms are used
  % rather than the generic algorithms used to infer other random variables
  % (such as Metropolis Hastings).

  % We define an example probabilistic programming language with this feature
  % (Quipp) and show how it can be used to write machine learning models
  % concisely.

  In an ordinary probabilistic programming language (such as Church),
  it is possible to treat parameters as random variables.  This
  would allow ordinary inference algorithms to infer parameters.  However,
  there are advantages of having unknown functions as a feature
  in the language.
  First, it is easier to
  write programs without knowing the details of different parameterized distributions.
  Second, the system can use specialized algorithms to infer parameters faster.

  % Inference in these models is performed using the expectation-maximization
  % algorithm, with alternating steps of inferring latent variables and optimizing
  % parameters.

  % - Motivation for Quipp
  %   - Explanation of "unknown functions"
  %   - Writing machine learning algorithms as probabilistic programs
  %   - Accessibility to non-experts
  %   - Comparison to existing probabilistic programming languages
  %     - In other languages, use random variables for parameters
  %     - Random variables slower because they are updated independently

  In the following, we first specify the syntax used to write Quipp programs,
  including the notation for unknown variables and functions.
  We describe the class of exponential family variables and functions that our system can learn,
  and present the expectation maximization algorithm used to learn them.
  We then demonstrate the expressiveness of our language, and the broad
  applicability of our algorithm, by writing some of the most common machine learning models
  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
  a neural net.

  \section{SYNTAX}

  Quipp is implemented as a library for webppl programs.  Webppl [TODO cite] is a probabilistic programming language
  implemented in Javascript   Quipp programs are written as webppl programs that have access to special functions.

  Here is an example of a Quipp program to cluster 2d points into 3 clusters:

  \begin{verbatim}
  dim = 2
  nclusters = 3
  PointType = Vector(dim, Double)
  ClusterType = Categorical(nclusters)

  get_point = rand_function(ClusterType, PointType)

  def sample():
    cluster = uniform_categorical(nclusters)
    return (cluster, get_point(cluster))
  \end{verbatim}

  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and one random function \texttt{get\_point}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
  the distribution over clusters is uniform, but we do not know the distribution
  of points in each cluster.
  We will fill in the
  \texttt{get\_point} function with a learned function that will take a random
  sample from the given cluster.

  \section{FAMILY OF DISTRIBUTIONS}

  For unknown functions, the family of random functions used is a kind
    of generalized linear model.  We assume that the result
    is distributed from some exponential family whose natural
    parameters are determined from the arguments:

    $$p_{\eta}(y | x) = \exp\left(\eta(x)^T \phi(y) - g(\eta(x))\right)$$

    Here, $\eta$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
    and $g$ is the log partition function.

    To determine $\eta$ as a function of $x$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.  Features are the same as sufficient
    statistics for the categorical distribution, but while both $X$ and $X^2$ are
    sufficient statistics for the Gaussian distribution, only $X$ is a feature.

    Let $\psi(x)$ be the features of $x$.  Then
    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for entries not in the first
    row corresponding to sufficient statistics of $y$ that are not features.  This ensures that only the
    natural parameters for features are affected by $x$.

  \section{INFERENCE}

    To infer both latent variables and parameters, we run the
    expectation maximization algorithm on the probabilistic model, iterating stages of
    estimating latent variables using Metropolis Hastings and inferring
    parameters using gradient descent.

    For the expectation step, we must estimate latent variable distributions given
    fixed values for the parameters.  To do this, we can run the Quipp program
    with random functions set to use these fixed parameter values to generate their results.
    We use the Metropolis Hastings algorithm to perform inference in this program,
    yielding traces.  Next, for each random function, we can find all calls
    to it in the trace to get the training data.

    For the maximization step, given samples from each random function, we set the
    parameters of the function to maximize the likelihood of the samples.  To do this we,
    we use gradient descent.

    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
    problem because the log probability function is concave:
    $$\log p_{\mathbf{N}}(y | x) = \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)$$

    This relies on the fact that $g$ is convex, but this is true in general for any exponential family distribution.
    Since the problem is convex, it is possible to use gradient descent to optimize the parameters.  Although
    the only exponential family distributions we use in this paper are the categorical and Gaussian distributions,
    we can use the same algorithms for other exponential families, such as the Poisson and gamma distributions.

  \section{EXAMPLES}

  \subsection{CLUSTERING}
{\small
\begin{verbatim}
dim = 2
nclusters = 3

PointType = Vector(dim, Double)
ClusterType = Categorical(nclusters)

get_cluster = rand_function(ClusterType)
get_point = rand_function(ClusterType, PointType)

def sample():
  cluster = get_cluster()
  return (cluster, get_point(cluster))
\end{verbatim}
}

In this example, we cluster 2d points into 3 different clusters.  Given a cluster, the distribution for a point is some independent Gaussian distribution.  This is similar to fuzzy c-means clustering.

We randomly generated parameters for this example 100 times, and each time took 100 samples and then ran 10 EM iterations.  The accuracy is defined as the maximum percentage of points assigned to the correct cluster, for any permutation of clusters.  On average, accuracy increased in each EM iteration, as shown it this graph:

\begin{center}
\includegraphics[scale=0.4]{cluster_accuracy.png}
\end{center}

\subsection{NAIVE BAYES}

{\small
\begin{verbatim}
nfeatures = 10
FeaturesType = Vector(nfeatures, Bool)

get_class = rand_function(Unit, Bool)

class_1_features = rand_function(Unit, FeaturesType)
class_2_features = rand_function(Unit, FeaturesType)

def sample():
  which_class = get_class()
  if which_class:
    features = class_1_features()
  else:
    features = class_2_features()
  return (which_class, features)
\end{verbatim}
}

The naive Bayes model is similar to the clustering model.  We have two classes and a feature
distribution for each.  Since each feature is boolean, we will learn
a different categorical distribution for each class.

(figure should show average classification accuracy)

  \subsection{FACTOR ANALYSIS}
{\small
\begin{verbatim}
num_components = 3
point_dim = 5

ComponentsType = Vector(num_components, Double)
PointType = Vector(point_dim, Double)

get_point = rand_function(ComponentsType, PointType)

def sample():
  components = [normal(0, 1)
                for i in range(num_components)]
  return (comenents, get_point(components))
\end{verbatim}
}

The factor analysis model is very similar to the clustering model.  The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.  This results in the model attempting to find each point as an affine function of a vector of standard normal values.

(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
  \subsection{HIDDEN MARKOV MODEL}
{\small
\begin{verbatim}
num_states = 5
num_observations = 3
sequence_length = 10

StateType = Categorical(num_states)
ObsType = Categorical(num_observations)

trans_fun = rand_function(StateType, StateType)
obs_fun = rand_function(StateType, ObsType)

def sample():
  states = [uniform(num_states)]
  for i in range(sequence_length - 1):
    states.append(trans_fun(states[-1]))
  return (states, [obs_fun(s) for s in states])
\end{verbatim}
}

In this example, we use the unknown function \texttt{trans\_fun} for state transitions and \texttt{obs\_fun} for observations.  This means that we will learn both the state transitions and the observation distribution.

\subsection{LATENT DIRICHLET ALLOCATION}
{\small
\begin{verbatim}
n_classes = 10
n_words = 100
n_words_per_document = 1000

ClassType = Categorical(n_classes)
WordType = Categorical(n_words)

class_to_word = rand_function(ClassType, WordType)

def sample():
    which_class = uniform_categorical(n_classes)
    words = [class_to_word(which_class)
             for i in range(n_words_per_document)]
    return (which_class, words)
\end{verbatim}
}

In this example, we use the unknown function \texttt{class\_to\_word} to map classes to word distributions.  Note that each column of the matrix of parameters for \texttt{class\_to\_word} will represent a categorical distribution over words, and there will be one column for each class.

(figure should show accuracy over time.  Accuracy can be measured as distance between the learned categorical distributions, for some permutation of classes)

\subsection{NEURAL NETWORK}

{\small
\begin{verbatim}
input_dim = 100
hidden_dim = 20
output_dim = 2

InputType = Categorical(input_dim)
HiddenType = Categorical(hidden_dim)
OutputType = Categorical(output_dim)

input_to_hidden = rand_function(InputType, HiddenType)
hidden_to_output = rand_function(HiddenType, OutputType)

def sample(input_layer):
  hidden_layer = input_to_hidden(input_layer)
  output_layer = hidden_to_output(hidden_layer)
  return ((), output_layer)
\end{verbatim}
}

\subsection{A MORE COMPLEX MODEL}

TODO: if there is time, we should put a more complex data science type example, where we add dependencies and show change in accuracy as we add/remove assumptions.

  \section{DISCUSSION}

  We have found that it is possible to write many useful machine learning models as Quipp programs and then use generic algorithms for inference.  Furthermore, performance is <???>.  This should make it much easier for non-experts to write useful machine learning models.

  In the future, it will be useful to support unbounded recursion in models (as in Church).  Furthermore, it will also be useful to expand the set of types supported.  It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.  Also, it will be useful to create a more usable interface to infer parameters and perform additional data pracessing given these parameters.

\end{document}

### 2015-05-25 src/expfam.js

@@ -86,6 +86,7 @@ var Double = {
 };
 
 function Categorical(n) {
+  assert(typeof n == 'number');
   return {
     name: 'Categorical(' + n + ')',
     sufStat: function(value) {
@@ -244,7 +245,8 @@ function gaussianMle(samples) {
   var ys = samples.map(function(samp) {
     return samp[2][0];
   });
-  // TODO: avoid creating weights matrix
+  // console.log('xss', xss);
+  // console.log('ys', ys);
   var A = opt.matMul(opt.transpose(xss), opt.elemMatProduct(weightsDiag, xss));
   var b = opt.matMulByVector(opt.transpose(xss), opt.elemProduct(weightsDiag, ys));
   var beta = opt.linSolve(A, b);
@@ -298,6 +300,32 @@ function mle(ef, samples, params) {
       paramsToVector(params)));
 }
 
+function defaultParameters(signature) {
+  var argTypes = signature[0];
+  var retType = signature[1];
+  var nfeatures = 0;
+  argTypes.forEach(function(at) {
+    nfeatures += featuresDim(at);
+  });
+  return {
+    base: retType.defaultNatParam,
+    weights: _.times(featuresDim(retType), function() {
+      return _.times(nfeatures, function() { return Math.random() * 20 - 10; });
+    })
+  };
+}
+
+var simpleSample = fromMonad(function(expfam, nps) {
+  var oldSample = global.sample;
+  global.sample = fromMonad(function(erp, params) {
+    return mreturn(erp.sample(params));
+  });
+  return mbind(expfam.sample, nps, function(res) {
+    global.sample = oldSample;
+    return mreturn(res);
+  });
+});
+
 module.exports = {
   dim: dim,
   featuresDim: featuresDim,
@@ -308,6 +336,8 @@ module.exports = {
   logProbability: logProbability,
   getNatParam: getNatParam,
   mle: mle,
-  paramsScoreFunction: paramsScoreFunction
+  paramsScoreFunction: paramsScoreFunction,
+  defaultParameters: defaultParameters,
+  simpleSample: simpleSample
 };
 

### 2015-05-25 src/mh.js

@@ -122,7 +122,6 @@ module.exports = function(env) {
       },
       exit: function(s, val) {
         env.coroutine = oldCoro;
-        console.log('recompute score', currScore);
         return k0(s0, [newTrace, currScore]);
       },
       factor: function(s, k, a, score) {

### 2015-05-25 src/param_inference.js

@@ -6,23 +6,60 @@ var expfam = require('./expfam');
 var util = require('./util');
 
 var mh = require('./mh')(webpplEnv);
+var pfais = require('./pfais')(webpplEnv);
 
 var mbind = util.mbind, mreturn = util.mreturn, mbindMethod = util.mbindMethod, fromMonad = util.fromMonad, mcurry = util.mcurry, mcurryMethod = util.mcurryMethod;
 
-function defaultParameters(signature) {
-  var argTypes = signature[0];
-  var retType = signature[1];
-  var nfeatures = 0;
-  argTypes.forEach(function(at) {
-    nfeatures += expfam.featuresDim(at);
-  });
-  return {
-    base: retType.defaultNatParam,
-    weights: _.times(expfam.featuresDim(retType), function() {
-      return _.times(nfeatures, function() { return Math.random() * 10 - 5; });
-    })
-  };
+function DataSampler() {
+  this.valueMap = {};
 }
+
+DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
+  var self = this;
+  var args = [].slice.call(arguments, 2, arguments.length);
+  if (typeof addr == 'function') {
+    return mbind(util.getAddress, function(a) {
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+    });
+  }
+  return mbind.apply(null, [rf].concat(args).concat([function(samp) {
+    self.valueMap[addr] = samp;
+    return mreturn(samp);
+  }]));
+});
+
+var sampleData = fromMonad(function(s) {
+  var sampler = new DataSampler();
+  global.observe = sampler.sampleOrCondition.bind(sampler);
+  return mbind(s, function(res) {
+    return mreturn(sampler.valueMap);
+  });
+});
+
+function DataConditioner(valueMap) {
+  this.valueMap = valueMap;
+}
+
+DataConditioner.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
+  var self = this;
+  var args = [].slice.call(arguments, 2, arguments.length);
+  if (typeof addr == 'function') {
+    return mbind(util.getAddress, function(a) {
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+    });
+  }
+  assert(addr in self.valueMap);
+  var ret = self.valueMap[addr];
[…]

### 2015-05-25 src/pfais.js

@@ -1,0 +1,257 @@
+////////////////////////////////////////////////////////////////////
+// Particle filtering
+//
+// Sequential importance re-sampling, which treats 'factor' calls as
+// the synchronization / intermediate distribution points.
+
+'use strict';
+
+var _ = require('underscore');
+var ad = require('./autodiff');
+var webpplUtil = require('webppl/src/util');
+var erp = require('webppl/src/erp');
+var expfam = require('./expfam');
+var util = require('./util');
+var mbind = util.mbind, mreturn = util.mreturn, mcurry = util.mcurry, fromMonad = util.fromMonad;
+
+
+module.exports = function(env) {
+
+  function copyParticle(particle) {
+    return {
+      continuation: particle.continuation,
+      weight: particle.weight,
+      value: particle.value,
+      store: _.clone(particle.store)
+    };
+  }
+
+  function erpRetType(erpp, params) {
+    if (erpp == erp.gaussianERP) {
+      return expfam.Double;
+    } else if (erpp == erp.bernoulliERP) {
+      return expfam.Categorical(2);
+    } else if (erpp == erp.randomIntegerERP) {
+      return expfam.Categorical(params[0]);
+    } else if (erpp == erp.discreteERP) {
+      return expfam.Categorical(params[0].length);
+    }
+    console.log('argh :/', erpp)
+    return null;
+  }
+
+  function ParticleFilter(s, k, a, wpplFn, numParticles, aisIters, strict) {
+
+    this.aisIters = aisIters;
+    this.aisRetTypes = [];
+    this.aisTrainingSamples = [];
+    this.aisParams = [];
+    this.aisIter = 0;
+    this.sampleIndex = 0;
+    this.particles = [];
+    this.particleIndex = 0;  // marks the active particle
+
+    // Create initial particles
+    var exitK = function(s) {
+      return wpplFn(s, env.exit, a);
+    };
+    for (var i = 0; i < numParticles; i++) {
+      var particle = {
+        continuation: exitK,
+        weight: 0,
+        value: undefined,
+        store: _.clone(s)
+      };
+      this.particles.push(particle);
+    }
+    // TODO clone store?
+    this.particlesAtLastFactor = this.particles.map(copyParticle);
+
+    this.strict = strict;
+    // Move old coroutine out of the way and install this as the current
+    // handler.
+    this.k = k;
+    this.oldCoroutine = env.coroutine;
+    env.coroutine = this;
+
+    this.oldStore = _.clone(s); // will be reinstated at the end
+  }
+
[…]

### 2015-05-25 src/quipp.js

@@ -19,6 +19,7 @@ module.exports = {
   factorScore: factorScore,
   inferParameters: param_inference.inferParameters,
   testParamInference: param_inference.testParamInference,
+  testParamInference2: param_inference.testParamInference2,
   Double: expfam.Double,
   Tuple: util.makeWpplFunction(expfam.Tuple),
   Vector: Vector,

### 2015-05-25 src/util.js

@@ -37,6 +37,20 @@ function catchStack(x) {
   }
 }
 
+function getAddress(s, k, a) {
+  return k(s, a);
+}
+
+function trampoline(f) {
+  return function(s, k, a) {
+    return function() {
+      return f(s, k, a)();
+    };
+  };
+}
+
+// 18008723377
+// 1800 traders
 
 function mbind(x) {
   assert(typeof x == 'function', x);
@@ -44,9 +58,9 @@ function mbind(x) {
   var extraArgs = args.slice(1, args.length - 1);
   var f = args[args.length - 1];
   assert(typeof f == 'function', f);
-  return function(s, k, a) {
+  return trampoline(function(s, k, a) {
     var self = this;
-    return catchStack(function() { return x.apply(self, [s, k2, a].concat(extraArgs)); });
+    return trampoline(catchStack(function() { return x.apply(self, [s, k2, a].concat(extraArgs)); }));
     function k2(s2, res) {
       return catchStack(function() {
         var nextf = f(res);
@@ -54,7 +68,7 @@ function mbind(x) {
         return nextf(s2, k, a);
       });
     }
-  }
+  });
 }
 
 function mbindMethod(x, method) {
@@ -85,21 +99,20 @@ function fromMonad(msg, f) {
   assert(typeof f == 'function');
   return function(s, k, a) {
     var args = [].slice.call(arguments, 3);
-    if (msg == null) {
-      return f.apply(this, args)(s, k, a);
-    } else {
-      return withStackMessage(msg, f.apply(this, args))(s, k, a);
-    }
+    // return function() {
+      if (msg == null) {
+        return f.apply(this, args)(s, k, a);
+      } else {
+        return withStackMessage(msg, f.apply(this, args))(s, k, a);
+      }
+    // };
   };
 }
 
 function mcurry(f) {
   assert(typeof f == 'function');
   var args = [].slice.call(arguments, 1);
-  return function(s, k, a) {
-    var extraArgs = [].slice.call(arguments, 3);
-    return f.apply(this, [s, k, a].concat(args).concat(extraArgs));
-  };
+  return mbind.apply(this, [f].concat(args).concat([mreturn]));
 }
 
 function mcurryMethod() {
@@ -213,6 +226,7 @@ function wpplMap(s, k, a, f, xs) {
 }
 
 module.exports = {
+  getAddress: getAddress,
   mbind: mbind,
   mbindMethod: mbindMethod,
   mreturn: mreturn,

### 2015-05-27 src/expfam.js
more examples working

@@ -74,6 +74,19 @@ var Double = {
       });
     });
   }),
+  randParams: fromMonad(function(nfeatures) {
+    return mbind(global.sample, erp.gaussianERP, [0, 5], function(mean) {
+      return mbind(global.sample, erp.gaussianERP, [0, 5], function(stdev) {
+        return mbind(util.replicateM, nfeatures, mcurry(global.sample, erp.gaussianERP, [0, 5]), function(coeffs) {
+          var variance = stdev*stdev;
+          return mreturn({
+            base: [mean / variance, -1 / (2 * variance)],
+            weights: [coeffs.map(function(coeff) { return coeff / variance; })]
+          });
+        });
+      });
+    });
+  }),
   featuresMask: [true, false],
   mle: function(samples, natParam) {
     return gaussianMle(samples);
@@ -145,6 +158,13 @@ function Tuple(types) {
     defaultNatParam: util.concat(_.pluck(types, 'defaultNatParam')),
     randNatParam: mbind(util.mapM, _.pluck(types, 'randNatParam'), fromMonad(function(x) { return x; }), function(res) {
       return mreturn(util.concat(res));
+    }),
+    randParams: fromMonad(function(nfeatures) {
+      return mbind(util.mapM, types, fromMonad(function(t) { return mcurry(t.randParams, nfeatures); }), function(res) {
+        console.log('res', res);
+        return mreturn({base: util.concat(_.pluck(res, 'base')),
+                        weights: util.concat(_.pluck(res, 'weights'))});
+      });
     }),
     featuresMask: util.concat(_.pluck(types, 'featuresMask')),
     mle: function(samples, params) {
@@ -310,7 +330,7 @@ function defaultParameters(signature) {
   return {
     base: retType.defaultNatParam,
     weights: _.times(featuresDim(retType), function() {
-      return _.times(nfeatures, function() { return Math.random() * 20 - 10; });
+      return _.times(nfeatures, function() { return Math.random() * 10 - 5; });
     })
   };
 }

### 2015-05-27 src/param_inference.js

@@ -68,10 +68,11 @@ var randParameters = fromMonad('randParameters', function(signature) {
   argTypes.forEach(function(at) {
     nfeatures += expfam.featuresDim(at);
   });
+  if (retType.randParams) return mcurry(retType.randParams, nfeatures);
   return mbind(retType.randNatParam, function(base) {
     assert(global.sample);
     return mbind(util.replicateM, expfam.featuresDim(retType),
-                 mcurry(util.replicateM, nfeatures, mcurry(global.sample, erp.gaussianERP, [0, 5])),
+                 mcurry(util.replicateM, nfeatures, mcurry(global.sample, erp.gaussianERP, [0, 2])),
                  function(weights) {
                    return mreturn({base: base, weights: weights});
                  });
@@ -87,6 +88,7 @@ function natParamWithArguments(argTypes, retType, params, argVals) {
 }
 
 function trainParameters(signature, calls, params) {
+  // console.log('calls', JSON.stringify(calls));
   var argTypes = signature[0];
   var retType = signature[1];
   var samps = calls.map(function(call) {
@@ -159,7 +161,7 @@ UnknownParametersModel.prototype.randSample = function(s, k, a, params) {
 
 UnknownParametersModel.prototype.logPartition = fromMonad(function(params) {
   var self = this;
-  var nparticles = 50;
+  var nparticles = 60;
   return mbind(global.ParticleFilterRejuv, self.getSamplerWithParameters(params), nparticles, 0, function(dist) {
     //return mbind(pfais.ParticleFilter, self.getSamplerWithParameters(params), nparticles, 5, function(dist2) {
       // console.log(dist.normalizationConstant, dist2.normalizationConstant);
@@ -178,7 +180,8 @@ UnknownParametersModel.prototype.formatParameters = function(paramsList) {
 UnknownParametersModel.prototype.stepParamsAndTrace = fromMonad(function(params, trace) {
   var self = this;
   // TODO: should this depend on something?
-  var numSamps = 2000;
+  // var numSamps = 2000;
+  var numSamps = 200;
   return mbind(mh.MH, self.getSamplerWithParameters(params), numSamps, trace, function(sampsDistrAndTrace) {
     console.log('got samples');
     var sampsDistr = sampsDistrAndTrace[0];
@@ -209,14 +212,16 @@ UnknownParametersModel.prototype.stepParamsAndTrace = fromMonad(function(params,
 
[…]

### 2015-05-27 src/param_inference.js
factor analysis seems to work

@@ -161,9 +161,9 @@ UnknownParametersModel.prototype.randSample = function(s, k, a, params) {
 
 UnknownParametersModel.prototype.logPartition = fromMonad(function(params) {
   var self = this;
-  var nparticles = 60;
+  var nparticles = 100;
   return mbind(global.ParticleFilterRejuv, self.getSamplerWithParameters(params), nparticles, 0, function(dist) {
-    //return mbind(pfais.ParticleFilter, self.getSamplerWithParameters(params), nparticles, 5, function(dist2) {
+    // return mbind(pfais.ParticleFilter, self.getSamplerWithParameters(params), nparticles, 5, function(dist) {
       // console.log(dist.normalizationConstant, dist2.normalizationConstant);
       return mreturn(dist.normalizationConstant);
     // });

### 2015-05-27 src/pfais.js

@@ -12,7 +12,37 @@ var webpplUtil = require('webppl/src/util');
 var erp = require('webppl/src/erp');
 var expfam = require('./expfam');
 var util = require('./util');
+var optimization = require('./optimization'); 
 var mbind = util.mbind, mreturn = util.mreturn, mcurry = util.mcurry, fromMonad = util.fromMonad;
+
+function objective(t, ef, samps, np) {
+  var unnormalizedLps = samps.map(function(samp) {
+    return ad.numDotProduct(t, np, samp[1].map(t.num));
+  });
+  var estG = ad.numLogSumExp(t, unnormalizedLps);
+  estG = t.sub(estG, t.num(Math.log(samps.length)));
+  var actualG = ef.g(t, np);
+  var totLp = t.num(0);
+  unnormalizedLps.forEach(function(lp, i) {
+    var thisLp = t.sub(t.sub(lp, estG), t.mul(t.num(0.01), actualG));
+    totLp = t.add(totLp, t.mul(t.num(samps[i][0]), thisLp));
+  });
+  return totLp;
+}
+
+function optNp(ef, samps, origNp) {
+  function scoreFn(t, np) {
+    return objective(t, ef, samps, np);
+  }
+  return optimization.gradientDescent(
+      function(np) {
+        return scoreFn(ad.standardNumType, np);
+      },
+      function(np) {
+        return ad.gradient(ad.standardNumType, scoreFn, np).grad;
+      },
+      origNp);
+}
 
 
 module.exports = function(env) {
@@ -28,13 +58,15 @@ module.exports = function(env) {
 
   function erpRetType(erpp, params) {
     if (erpp == erp.gaussianERP) {
-      return expfam.Double;
+      var mean = params[0];
+      var variance = params[1];
+      return [expfam.Double, [mean / variance, 1 / (2 * variance)]];
     } else if (erpp == erp.bernoulliERP) {
-      return expfam.Categorical(2);
+      return [expfam.Categorical(2), null];
     } else if (erpp == erp.randomIntegerERP) {
-      return expfam.Categorical(params[0]);
+      return [expfam.Categorical(params[0]), _.times(params[0] - 1, function() { return 0; })];
     } else if (erpp == erp.discreteERP) {
-      return expfam.Categorical(params[0].length);
+      return [expfam.Categorical(params[0].length), null];
     }
     console.log('argh :/', erpp)
     return null;
[…]

### 2015-05-28 paper/paper.tex
som stuff

@@
 \And
+Andreas Stuhlm\"uller \\
+Department of Brain and Cognitive Sciences \\
+MIT \\
+Cambridge, MA 02139 \\
+\texttt{andreas@stuhlmueller.org}
+\And
 Noah Goodman \\
-Department of ?? \\
+Department of Psychology \\
 Stanford University \\
 Stanford, CA 94305
-\texttt{...}
-\And
+\texttt{ngoodman@stanford.edu}
 }
@@
 
-  \section{INTRODUCTION}
+  \section{Introduction}
 
@@
 
-  \section{SYNTAX}
+  \section{Syntax}
 
   Quipp is implemented as a library for webppl programs.  Webppl [TODO cite] is a probabilistic programming language
-  implemented in Javascript   Quipp programs are written as webppl programs that have access to special functions.
+  implemented in Javascript. Quipp programs are written as webppl programs that have access to special functions.
 
@@
   \begin{verbatim}
-  dim = 2
-  nclusters = 3
-  PointType = Vector(dim, Double)
-  ClusterType = Categorical(nclusters)
+var nclusters = 3;
+var dim = 2;
 
-  get_point = rand_function(ClusterType, PointType)
+var n = 100;
 
-  def sample():
-    cluster = uniform_categorical(nclusters)
-    return (cluster, get_point(cluster))
+testParamInference(function(randFunction) {
+  var pointType = Vector(dim, Double);
+  var getPoint = randFunction(Categorical(nclusters), pointType);
+  return function() {
+    return repeat(n, function(i) {
+      var cluster = randomInteger(nclusters);
+      return observe('point' + i, getPoint, cluster);
+    });
+  };
+});
   \end{verbatim}
 
-  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
-  and one random function \texttt{get\_point}).
+  It is written as a generative model producing the observations.  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
+  and one random function \texttt{getPoint}).
   Type annotations are necessary for random functions.  The type
@@
   We will fill in the
-  \texttt{get\_point} function with a learned function that will take a random
+  \texttt{getPoint} function with a learned function that will take a random
   sample from the given cluster.
 
-  \section{FAMILY OF DISTRIBUTIONS}
+  \section{Family of distributions}
 
   For unknown functions, the family of random functions used is a kind
-    of generalized linear model.  We assume that the result
-    is distributed from some exponential family whose natural
+    of generalized linear model.  We assume that the distribution
+    of the function's return value
+    is some exponential family whose natural
     parameters are determined from the arguments:
@@
 
-    Here, $\eta$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
+    Here, $\eta(x)$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
     and $g$ is the log partition function.
 
-    To determine $\eta$ as a function of $x$, we label
+    To determine $\eta(x)$, we label
     some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
@@
 
-    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for entries not in the first
-    row corresponding to sufficient statistics of $y$ that are not features.  This ensures that only the
-    natural parameters for features are affected by $x$.
+    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
+    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
+    This ensures that only the
+    natural parameters that are features of $y$ are affected by $x$.
 
-  \section{INFERENCE}
+  \section{Inference}
 
@@
 
-  \section{EXAMPLES}
+  \section{Evaluation}
 
-  \subsection{CLUSTERING}
+    To evaluate performance, for each model, we:
+    \begin{itemize}
+      \item
+        Randomly generate parameters $\theta$
+      \item
+        Generate datasets $x_{train}, x_{test}$ using $\theta$
+      \item
+        Estimate $\log P(x_{test} | \theta)$
+      \item
+        Use the EM algorithm to infer approximate parameters $\hat{\theta}$ from $x_{train}$
+      \item
+        Estimate $\log P(x_{test} | \hat{\theta})$ and compare to $\log P(x_{test} | \theta)$
+    \end{itemize}
+    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
+    We use the Sequential Monte Carlo algorithm for this.  Between observations,
+
+  \section{Examples}
+
+  \subsection{Clustering}
 {\small
 \begin{verbatim}
-dim = 2
-nclusters = 3
+var nclusters = 3;
+var dim = 2;
 
-PointType = Vector(dim, Double)
-ClusterType = Categorical(nclusters)
+var n = 100;
 
-get_cluster = rand_function(ClusterType)
-get_point = rand_function(ClusterType, PointType)
-
-def sample():
-  cluster = get_cluster()
-  return (cluster, get_point(cluster))
+testParamInference(function(randFunction) {
+  var pointType = Vector(dim, Double);
+  var getPoint = randFunction(Categorical(nclusters), pointType);
+  return function() {
+    return repeat(n, function(i) {
+      var cluster = randomInteger(nclusters);
+      return observe('point' + i, getPoint, cluster);
+    });
+  };
+});
 \end{verbatim}
@@
 
-\subsection{NAIVE BAYES}
+\subsection{Naive Bayes}
 
@@
 
-  \subsection{FACTOR ANALYSIS}
+  \subsection{Factor analysis}
 {\small
@@
 (figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
-  \subsection{HIDDEN MARKOV MODEL}
+  \subsection{Hidden Markov model}
 {\small
@@
 
-\subsection{LATENT DIRICHLET ALLOCATION}
+\subsection{Latent Dirichlet allocation}
 {\small
@@
 
-\subsection{NEURAL NETWORK}
+\subsection{Neural network}
 
@@
 
-\subsection{A MORE COMPLEX MODEL}
+\subsection{A more complex model}
 
@@
 
-  \section{DISCUSSION}
+  \section{Discussion}
 
@@
 
-  In the future, it will be useful to support unbounded recursion in models (as in Church).
-Furthermore, it will also be useful to expand the set of types supported.
+  In the future, it will be useful to expand the set of types supported.
 It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.

### 2015-05-28 src/param_inference.js

@@ -256,7 +256,7 @@ var generateRandData = fromMonad(function(fun, params) {
   });
 });
 
-var testParamInference = fromMonad(function(fun) {
+var testParamInferenceSplit = fromMonad(function(fun) {
   console.log('testParamInference');
   return mbind(unknownParametersModel, fun, function(upmWrong) {
     console.log('upmwrong', upmWrong.signatures);
@@ -301,7 +301,7 @@ var testParamInference = fromMonad(function(fun) {
   });
 });
 
-var testParamInference2 = fromMonad(function(fun) {
+var testParamInference = fromMonad(function(fun) {
   var fun2 = fromMonad(function(randFunction) {
     return mbind(fun, randFunction, function(innerSampler) {
       var gendata = mcurry(sampleData, innerSampler);
@@ -311,12 +311,12 @@ var testParamInference2 = fromMonad(function(fun) {
       return mreturn([gendata, condata]);
     });
   });
-  return mcurry(testParamInference, fun2);
+  return mcurry(testParamInferenceSplit, fun2);
 });
 
 
 module.exports = {
   inferParameters: inferParameters,
   testParamInference: testParamInference,
-  testParamInference2: testParamInference2
+  testParamInferenceSplit: testParamInferenceSplit
 };

### 2015-05-28 src/quipp.js

@@ -19,7 +19,7 @@ module.exports = {
   factorScore: factorScore,
   inferParameters: param_inference.inferParameters,
   testParamInference: param_inference.testParamInference,
-  testParamInference2: param_inference.testParamInference2,
+  testParamInferenceSplit: param_inference.testParamInferenceSplit,
   Double: expfam.Double,
   Tuple: util.makeWpplFunction(expfam.Tuple),
   Vector: Vector,

### 2015-05-30 paper/paper.tex
paper

@@
   Here is an example of a Quipp program to cluster 2d points into 3 clusters:
-
   \begin{verbatim}
@@
   return function() {
-    return repeat(n, function(i) {
+    repeat(n, function(i) {
       var cluster = randomInteger(nclusters);
-      return observe('point' + i, getPoint, cluster);
+      observe('point' + i, getPoint, cluster);
     });
@@
 
-  It is written as a generative model producing the observations.  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
-  and one random function \texttt{getPoint}).
-  Type annotations are necessary for random functions.  The type
-  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
-  and the type \texttt{Categorical(3)} expresses the fact that there are 3
-  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
-  the distribution over clusters is uniform, but we do not know the distribution
-  of points in each cluster.
-  We will fill in the
+  It is written as a generative model producing the observations.  Notice that
+  we declared two types (\texttt{PointType} and \texttt{ClusterType}) and one
+  random function \texttt{getPoint}).  Type annotations are necessary for random
+  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
+  points are 2-dimensional, and the type \texttt{Categorical(3)} expresses the
+  fact that there are 3 possible clusters (so a cluster may be either 0, 1, or
+  2).    We assume the distribution over clusters is uniform, but we do not know
+  the distribution of points in each cluster.  We will fill in the
   \texttt{getPoint} function with a learned function that will take a random
-  sample from the given cluster.
+  sample from the given cluster.  The \texttt{observe} function allows us to
+  observe data; here it says that the observation named \texttt{point3} is equal
+  to the result of calling \texttt{getPoint(cluster)}, where \texttt{cluster} is
+  cluster number 3.
+
+  To demonstrate, let us run this example on a dataset consisting of 150 points (TODO cite).  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
+  In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
+  split the data into 2 clusters with equal variance:
+  \begin{center}
+    \includegraphics[scale=0.5]{../plots/irisclusters_orig.png}
+  \end{center}
+
+  The first cluster is at (6.3, 5.0) and the second is at (5.0, 1.6).  They both
+  have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
+  We could use these parameters to fill in the generative model:
+  \begin{verbatim}
+    return repeat(n, function(i) {
+      var cluster = randomInteger(nclusters);
+      return [gaussian(cluster == 0 ? 6.3 : 5.0, 0.54),
+              gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
+    });
+  \end{verbatim}
+  This model is estimated to assign probability density $e^{-393}$ to the data, or 0.0728 per point.
+
+  Note that, because the two clusters are forced to have the same variance, they do not fit the data well,
+  since the data has a different shape in each location.  To fix this problem, we can substitute the following
+  model:
+  \begin{verbatim}
+testParamInference(function() {
+  var pointType = Vector(dim, Double);
+  var getPointFunctions = repeat(nclusters, function(i) {
+    return randFunction(pointType);
+  });
+  return function() {
+    return repeat(n, function(i) {
+      var cluster = randomInteger(nclusters);
+      return observe('point' + i, getPointFunctions[cluster]);
+    });
+  };
+});
+  \end{verbatim}
+  Using this model, we get the following clusters:
+  \begin{center}
+    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
+  \end{center}
+  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, which means it
+  fits the data better.
 
@@
 \begin{verbatim}
-nfeatures = 10
-FeaturesType = Vector(nfeatures, Bool)
+var nfeatures = 10;
+var featuresType = Vector(nfeatures, Bool);
+var nclasses = 2;
+var n = 50;
 
-get_class = rand_function(Unit, Bool)
+testParamInference(function(randFunction) {
+  var class0Features = randFunction(featuresType);
+  var class1Features = randFunction(featuresType);
 
-class_1_features = rand_function(Unit, FeaturesType)
-class_2_features = rand_function(Unit, FeaturesType)
-
-def sample():
-  which_class = get_class()
-  if which_class:
-    features = class_1_features()
-  else:
-    features = class_2_features()
-  return (which_class, features)
+  var getFeatures = randFunction(Categorical(nclusters), pointType);
+  return function() {
+    repeat(n, function(i) {
+      var whichClass = randomInteger(2);
+      if (whichClass == 0) {
+        observe('features' + i, class0Features);
+      } else {
+        observe('features' + i, class1Features);
+      }
+    });
+  };
+});
 \end{verbatim}
@@
 \begin{verbatim}
-num_components = 3
-point_dim = 5
 
-ComponentsType = Vector(num_components, Double)
-PointType = Vector(point_dim, Double)
+var nfactors = 2;
+var dim = 4;
 
-get_point = rand_function(ComponentsType, PointType)
+var n = 50;
 
-def sample():
-  components = [normal(0, 1)
-                for i in range(num_components)]
-  return (comenents, get_point(components))
+testParamInference(function(randFunction) {
+  var factorType = Vector(nfactors, Double);
+  var pointType = Vector(dim, Double);
+  var getPoint = randFunction(factorType, pointType);
+  var getFactors = function() {
+    return repeat(nfactors, function() { return gaussian(0, 1); });
+  };
+  return function() {
+    return repeat(n, function(i) {
+      var factors = getFactors();
+      return observe('point' + i, getPoint, factors);
+    });
+  };
+});
 \end{verbatim}
@@
 
-(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
   \subsection{Hidden Markov model}
@@
 \begin{verbatim}
-num_states = 5
-num_observations = 3
-sequence_length = 10
 
-StateType = Categorical(num_states)
-ObsType = Categorical(num_observations)
+var nstates = 2;
+var nobs = 3;
 
-trans_fun = rand_function(StateType, StateType)
-obs_fun = rand_function(StateType, ObsType)
+var chainLength = 8;
 
-def sample():
-  states = [uniform(num_states)]
-  for i in range(sequence_length - 1):
-    states.append(trans_fun(states[-1]))
-  return (states, [obs_fun(s) for s in states])
+var n = 30;
+
+testParamInference(function(randFunction) {
+  var stateType = Categorical(nstates);
+  var obsType = Categorical(nobs);
+  var transFun = randFunction(stateType, stateType);
+  var obsFun = randFunction(stateType, obsType);
+
+  var observeStates = function(sampIndex, startState, i) {
+    if (i == chainLength) {
+      return [];
+    } else {
+      observe('obs' + sampIndex + '_' + i, obsFun, startState);
+      return [startState].concat(observeStates(sampIndex, transFun(startState), i+1));
+    }
+  };
+  return function() {
+    return repeat(n, function(sampIndex) {
+      return observeStates(sampIndex, randomInteger(nstates), 0);
+    });
+  };
+});
 \end{verbatim}
@@
 
-In this example, we use the unknown function \texttt{trans\_fun} for state transitions and \texttt{obs\_fun} for observations.
+In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.
 This means that we will learn both the state transitions and the observation distribution.
@@
 \begin{verbatim}
-n_classes = 10
-n_words = 100
-n_words_per_document = 1000
+var nClasses = 2;
+var nWords = 100;
+var maxWordsPerDocument = 1000;
+var nDocuments = 20;
 
-ClassType = Categorical(n_classes)
-WordType = Categorical(n_words)
+var classType = Categorical(nClasses);
+var wordType = Categorical(nWords);
 
-class_to_word = rand_function(ClassType, WordType)
-
-def sample():
-    which_class = uniform_categorical(n_classes)
-    words = [class_to_word(which_class)
-             for i in range(n_words_per_document)]
-    return (which_class, words)
+testParamInference(function(randFunction) {
+  var classToWord = randFunction(ClassType, WordType);
+  return function() {
+    return repeat(nDocuments, function(docIndex) {
+      var whichClass = randomInteger(nClasses);
+      var nWords = observe('len' + docIndex, randomInteger, maxWordsPerDocument);
+      repeat(nWordsPerDocument, function(wordIndex) {
+        observe('word' + doc + '_' + wordIndex, classToWord, whichClass);
+      });
+      return whichClass;
+    });
+  };
+});
 \end{verbatim}
@@
 
-In this example, we use the unknown function \texttt{class\_to\_word} to map classes to word distributions.
-Note that each column of the matrix of parameters for \texttt{class\_to\_word} will represent a categorical distribution over words, and there will be one column for each class.
+In this example, we use the unknown function \texttt{classToWord} to map classes to word distributions.
+Note that each column of the matrix of parameters for \texttt{classToWord} will represent a categorical distribution over words, and there will be one column for each class.
 
@@
 \begin{verbatim}
-input_dim = 100
-hidden_dim = 20
-output_dim = 2
+var inputDim = 100;
+var hiddenDim = 20;
+var outputDim = 2;
 
-InputType = Categorical(input_dim)
-HiddenType = Categorical(hidden_dim)
-OutputType = Categorical(output_dim)
+var inputType = Categorical(inputDim);
+var hiddenType = Categorical(hiddenDim);
+var outputType = Categorical(outputDim);
 
-input_to_hidden = rand_function(InputType, HiddenType)
-hidden_to_output = rand_function(HiddenType, OutputType)
+var inputs = [...];
 
-def sample(input_layer):
-  hidden_layer = input_to_hidden(input_layer)
-  output_layer = hidden_to_output(hidden_layer)
-  return ((), output_layer)
+testParamInference(function(randFunction) {
+  var inputToHidden = randFunction(inputType, hiddenType);
+  var hiddenToOutput = randFunction(hiddenType, outputType);
+
+  return function() {
+    return repeat(inputs.length, function(sampIndex) {
+      var hiddenLayer = inputToHidden(inputs[sampIndex]);
+      observe('output' + sampIndex, hiddenToOutput, hiddenLayer);
+    });
+  };
+});
 \end{verbatim}

### 2015-05-30 src/expfam.js

@@ -161,7 +161,6 @@ function Tuple(types) {
     }),
     randParams: fromMonad(function(nfeatures) {
       return mbind(util.mapM, types, fromMonad(function(t) { return mcurry(t.randParams, nfeatures); }), function(res) {
-        console.log('res', res);
         return mreturn({base: util.concat(_.pluck(res, 'base')),
                         weights: util.concat(_.pluck(res, 'weights'))});
       });

### 2015-05-30 src/param_inference.js

@@ -161,7 +161,7 @@ UnknownParametersModel.prototype.randSample = function(s, k, a, params) {
 
 UnknownParametersModel.prototype.logPartition = fromMonad(function(params) {
   var self = this;
-  var nparticles = 100;
+  var nparticles = 10;
   return mbind(global.ParticleFilterRejuv, self.getSamplerWithParameters(params), nparticles, 0, function(dist) {
     // return mbind(pfais.ParticleFilter, self.getSamplerWithParameters(params), nparticles, 5, function(dist) {
       // console.log(dist.normalizationConstant, dist2.normalizationConstant);
@@ -239,7 +239,15 @@ var printReducer = fromMonad(function(params, trace, rest) {
 
 var inferParameters = fromMonad(function(fun) {
   return mbind(unknownParametersModel, fun, function(upm) {
-    return mcurryMethod(model, 'inferParameters', printReducer);
+    var reducer = fromMonad(function(params, trace, rest) {
+      console.log('yay', JSON.stringify(upm.formatParameters(params)));
+      console.log('also', JSON.stringify(params));
+      return mbindMethod(upm, 'logPartition', params, function(lp) {
+        console.log('score', lp);
+        return rest;
+      });
+    });
+    return mcurryMethod(upm, 'inferParameters', reducer);
   });
 });
 
@@ -283,11 +291,14 @@ var testParamInferenceSplit = fromMonad(function(fun) {
             return mbind(unknownParametersModel, innerSamplerForData(testData), function(upmTest) {
               console.log('upmTest');
               return mbindMethod(upmTest, 'logPartition', origParams, function(origLp) {
-                console.log('orig lp', origLp);
+                console.log('@', origLp);
+                var niters = 0;
                 var reducer = fromMonad(function(infParams, trace, rest) {
                   console.log('params', JSON.stringify(upmWrong.formatParameters(infParams)));
                   return mbindMethod(upmTest, 'logPartition', infParams, function(lp) {
-                    console.log('inf lp', lp);
+                    console.log('#', lp);
+                    niters += 1;
+                    if (niters == 3) process.exit()
                     return rest;
                   });
                 });

### 2015-06-01 getdata.py
getting plots

@@ -1,0 +1,43 @@
+import sys
+import subprocess
+import sys
+import json
+
+
+nparallel = 4
+niters = 5
+
+program_name = sys.argv[1]
+
+
+def append_result(line):
+  with open('plots/' + program_name + '.data', 'a+') as f:
+    f.write(line + '\n')
+
+# for run in range(nruns):
+#   output = subprocess.check_output(['./runquipp', 'examples/' + sys.argv[1] + '.wppl', '--iters', str(niters)])
+#   last_line = output.split('\n')[-2]
+#   print last_line
+#   run_results.append(json.loads(last_line))
+
+while True:
+  popens = []
+  for run in range(nparallel):
+    p = subprocess.Popen(['./runquipp', 'examples/' + program_name + '.wppl', '--iters', str(niters)],
+                         stdout=subprocess.PIPE, stderr=subprocess.PIPE)
+    popens.append(p)
+
+  for p in popens:
+    p.wait()
+    output = p.stdout.read()
+    error = p.stderr.read()
+    last_line = output.split('\n')[-2]
+    if 'results!' in last_line: 
+      append_result(last_line)
+    else:
+      print '-----------------'
+      print output
+      print '*****************'
+      print error
+      print '+++++++++++++++++'
+    print last_line

### 2015-06-01 plotaccuracy.py

@@ -1,0 +1,44 @@
+from pylab import *
+import sys
+import sys
+import json
+import subprocess
+
+
+niters = 30
+
+program_name = sys.argv[1]
+
+run_results = []
+for line in open('plots/' + program_name + '.data'):
+  run_results.append(json.loads(line))
+
+
+def concat(lst):
+  res = []
+  for l in lst:
+    res.extend(l)
+  return res
+
+ymax = 0
+ymin = 0
+for run in run_results:
+  origLp = run['origLp']
+  infLps = run['infLps'][1:]
+  regrets = [origLp - i for i in infLps]
+  regrets = [min(r, 1000) for r in regrets]
+  regrets = [max(r, -100) for r in regrets]
+  print 'regrets', regrets
+  ymax = max(ymax, max(regrets))
+  ymin = min(ymin, min(regrets))
+  plot(regrets)
+  # plot([0, len(infLps) - 1], [origLp, origLp])
+
+axis(ymin=ymin, ymax=ymax)
+
+plot_name = 'plots/accuracy_' + program_name + '.png'
+
+savefig(plot_name)
+
+subprocess.call(['eog', plot_name])
+

### 2015-06-01 src/expfam.js

@@ -117,7 +117,15 @@ function Categorical(n) {
     },
     defaultNatParam: _.times(n-1, function() { return 0.0; }),
     randNatParam: fromMonad(function() {
-      return mcurry(util.replicateM, n-1, mcurry(global.sample, erp.gaussianERP, [0, 5]));
+      return mcurry(util.replicateM, n-1, mcurry(global.sample, erp.gaussianERP, [0, 1]));
+    }),
+    randParams: fromMonad(function(nfeatures) {
+      var getRow = mcurry(util.replicateM, n-1, mcurry(global.sample, erp.gaussianERP, [0, 1]));
+      return mbind(getRow, function(base) {
+        return mbind(util.replicateM, nfeatures, getRow, function(weights) {
+          return mreturn({base: base, weights: opt.transpose(weights)});
+        });
+      });
     }),
     featuresMask: _.times(n-1, function() { return true; }),
     formatParams: function(params) { return params; }

### 2015-06-01 src/mh.js

@@ -179,7 +179,6 @@ module.exports = function(env) {
 
       return this.sample(_.clone(regen.store), regen.k, regen.name, regen.erp, regen.params, true);
     } else {
-      console.log('d', totDelta, this.currScore);
       var dist = erp.makeMarginalERP(this.returnHist);
 
       // Reinstate previous coroutine:

### 2015-06-01 src/param_inference.js

@@ -10,6 +10,9 @@ var pfais = require('./pfais')(webpplEnv);
 
 var mbind = util.mbind, mreturn = util.mreturn, mbindMethod = util.mbindMethod, fromMonad = util.fromMonad, mcurry = util.mcurry, mcurryMethod = util.mcurryMethod;
 
+var argv = require('yargs').argv;
+console.log('argv', argv);
+
 function DataSampler() {
   this.valueMap = {};
 }
@@ -22,6 +25,14 @@ DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
       return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
     });
   }
+
+  if (rf.constructor == erp.ERP) {
+    return mbind(global.sample, rf, args, function(samp) {
+      self.valueMap[addr] = samp;
+      return mreturn(samp);
+    });
+  }
+
   return mbind.apply(null, [rf].concat(args).concat([function(samp) {
     self.valueMap[addr] = samp;
     return mreturn(samp);
@@ -48,8 +59,16 @@ DataConditioner.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
       return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
     });
   }
+
   assert(addr in self.valueMap);
   var ret = self.valueMap[addr];
+
+  if (rf.constructor == erp.ERP) {
+    return mbind(global.factor, rf.score(args, ret), function() {
+      return mreturn(ret);
+    });
+  }
+
   return mcurry.apply(null, [rf.factorScore].concat(args).concat([ret]));
 });
 
@@ -161,12 +180,14 @@ UnknownParametersModel.prototype.randSample = function(s, k, a, params) {
 
 UnknownParametersModel.prototype.logPartition = fromMonad(function(params) {
   var self = this;
-  var nparticles = 10;
+  var nparticles = 30;
   return mbind(global.ParticleFilterRejuv, self.getSamplerWithParameters(params), nparticles, 0, function(dist) {
+  return mbind(global.ParticleFilterRejuv, self.getSamplerWithParameters(params), nparticles, 0, function(dist2) {
     // return mbind(pfais.ParticleFilter, self.getSamplerWithParameters(params), nparticles, 5, function(dist) {
       // console.log(dist.normalizationConstant, dist2.normalizationConstant);
[…]

### 2015-06-05 paper/paper.tex
paper progress

@@
 \documentclass{article}
+\usepackage{cite}
 \usepackage{amsmath}
+\usepackage{hyperref}
 \usepackage{amsfonts}
@@
 
-  Quipp is implemented as a library for webppl programs.  Webppl [TODO cite] is a probabilistic programming language
-  implemented in Javascript. Quipp programs are written as webppl programs that have access to special functions.
+  Quipp is implemented as a library for webppl programs.  Webppl \cite{dippl} is a probabilistic programming language
+  that is similar to Javascript but also contains features for generating random values, conditioning on values,
+  and estimating expectations. Quipp programs are written as webppl programs that have access to additional special functions.
 
-  Here is an example of a Quipp program to cluster 2d points into 3 clusters:
+  Here is an example of a Quipp program to cluster 2d points into 2 clusters:
   \begin{verbatim}
-var nclusters = 3;
-var dim = 2;
-
 var n = 100;
 
-testParamInference(function(randFunction) {
-  var pointType = Vector(dim, Double);
-  var getPoint = randFunction(Categorical(nclusters), pointType);
+var Cluster = Categorical(2);
+var Point = Vector(2, Double);
+var getPoint = randFunction(Categorical(2), pointType);
 return function() {
   repeat(n, function(i) {
-      var cluster = randomInteger(nclusters);
-      observe('point' + i, getPoint, cluster);
+    var cluster = randomValue(Cluster);
+    observe(getPoint, cluster);
   });
 };
-});
   \end{verbatim}
 
-  It is written as a generative model producing the observations.  Notice that
-  we declared two types (\texttt{PointType} and \texttt{ClusterType}) and one
+  It is written as a generative model for the observations.  Notice that
+  we declared two types (\texttt{Point} and \texttt{Cluster}) and one
   random function \texttt{getPoint}).  Type annotations are necessary for random
   functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
-  points are 2-dimensional, and the type \texttt{Categorical(3)} expresses the
-  fact that there are 3 possible clusters (so a cluster may be either 0, 1, or
-  2).    We assume the distribution over clusters is uniform, but we do not know
-  the distribution of points in each cluster.  We will fill in the
+  points are 2-dimensional, and the type \texttt{Categorical(2)} expresses the
+  fact that there are 2 possible clusters (so a cluster may be either 0 or 1).
+  We use the \texttt{randomValue} function to generate a random uniform cluster. We do not know
+  the distribution of points in each cluster, represented by the fact that \texttt{getPoint}
+  is an unknown random function.  We will fill in the
   \texttt{getPoint} function with a learned function that will take a random
@@
   since the data has a different shape in each location.  To fix this problem, we can substitute the following
-  model:
+  model, which uses a separate random function for each cluster:
   \begin{verbatim}
-testParamInference(function() {
-  var pointType = Vector(dim, Double);
-  var getPointFunctions = repeat(nclusters, function(i) {
-    return randFunction(pointType);
-  });
-  return function() {
+var Cluster = Categorical(2);
+var Point = Vector(2, Double);
+var getPointFunctions = [randFunction(Point), randFunction(Point)];
 return repeat(n, function(i) {
-      var cluster = randomInteger(nclusters);
-      return observe('point' + i, getPointFunctions[cluster]);
-    });
-  };
+  var cluster = randomValue(Cluster);
+  observe('point' + i, getPointFunctions[cluster]);
 });
@@
   \end{center}
-  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, which means it
-  fits the data better.
+  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, so it assigns significantly higher likelihood to the data.
+
+  Note that the clusters are still restricted to have $x$ and $y$ be independent.  To allow these coordinates
+  to be correlated within a cluster, we can use the following model:
+
+\begin{verbatim}
+
+var Cluster = Categorical(2);
+var getX = [randFunction(Double), randFunction(Double)];
+var getY = [randFunction(Double, Double), randFunction(Double, Double)];
+
+return function() {
+  return repeat(n, function(i) {
+    var cluster = randomInteger(nclusters);
+    var x = observe(getX[cluster]);
+    var y = observe(getY[cluster], x);
+  });
+};
+\end{verbatim}
+
+  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend on the $x$ coordinate).
+This allows $x$ and $y$ to be correlated, improving fit:
+
+  \begin{center}
+    \includegraphics[scale=0.5]{../plots/irisclusters_dep.png}
+  \end{center}
+
+  This model is estimated to assign probability $e^{-269}$ to the data, or 0.166 per point, a large improvement from the previous model.
 
@@
     of generalized linear model.  We assume that the distribution
-    of the function's return value
+    of the function's return value $y$
     is some exponential family whose natural
-    parameters are determined from the arguments:
+    parameters are determined from the arguments $x$:
 
@@
     To determine $\eta(x)$, we label
-    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
+    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
+    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
+    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
+    and features are both $[X = 1], [X = 2] ..., [X=n - 1]$.
+    The natural
     parameters corresponding to non-features are constant, while natural
     parameters corresponding to features are determined as an affine
-    function of the features of the arguments.  Features are the same as sufficient
-    statistics for the categorical distribution, but while both $X$ and $X^2$ are
-    sufficient statistics for the Gaussian distribution, only $X$ is a feature.
+    function of the features of the arguments.
 
     Let $\psi(x)$ be the features of $x$.  Then
+  $$\eta(x) = \mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}$$
     $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$
@@
     natural parameters that are features of $y$ are affected by $x$.
+
+    \begin{tabular}{| c | c | c | c|}
+      \hline
+      T & $\phi_T(X)$ & $\psi_T(X)$ & Random function class \\
+
+      \hline
+    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & Linear regression with Gaussian noise \\
+
+      \hline
+    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $n$-class logistic regression \\
+
+      \hline
+    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
+                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
+      Independent prediction of each component
+      \\
+      \hline
+    \end{tabular}
 
@@
 
-testParamInference(function(randFunction) {
 var pointType = Vector(dim, Double);
@@
 };
-});
 \end{verbatim}
@@
 
-testParamInference(function(randFunction) {
 var class0Features = randFunction(featuresType);
@@
 
-testParamInference(function(randFunction) {
 var factorType = Vector(nfactors, Double);
@@
 
-testParamInference(function(randFunction) {
 var stateType = Categorical(nstates);
@@
 };
+
 return function() {
@@
 
-testParamInference(function(randFunction) {
 var classToWord = randFunction(ClassType, WordType);
@@
 
+\bibliography{bibliography}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2015-06-05 getdata.py

@@ -5,9 +5,15 @@ import json
 
 
 nparallel = 4
-niters = 5
 
 program_name = sys.argv[1]
+
+niters = 30
+if program_name == 'factor_analysis':
+  niters = 12
+if program_name == 'hmm':
+  niters = 10
+
 
 
 def append_result(line):

### 2015-06-05 plotaccuracy.py

@@ -5,13 +5,13 @@ import json
 import subprocess
 
 
-niters = 30
-
 program_name = sys.argv[1]
 
 run_results = []
 for line in open('plots/' + program_name + '.data'):
   run_results.append(json.loads(line))
+
+print 'len', len(run_results)
 
 
 def concat(lst):
@@ -20,21 +20,48 @@ def concat(lst):
     res.extend(l)
   return res
 
-ymax = 0
-ymin = 0
-for run in run_results:
-  origLp = run['origLp']
-  infLps = run['infLps'][1:]
-  regrets = [origLp - i for i in infLps]
-  regrets = [min(r, 1000) for r in regrets]
-  regrets = [max(r, -100) for r in regrets]
-  print 'regrets', regrets
-  ymax = max(ymax, max(regrets))
-  ymin = min(ymin, min(regrets))
-  plot(regrets)
-  # plot([0, len(infLps) - 1], [origLp, origLp])
+# ymax = 0
+# ymin = 0
+# for run in run_results:
+#   origLp = run['origLp']
+#   infLps = run['infLps'][1:]
+#   regrets = [origLp - i for i in infLps]
+#   regrets = [min(r, 1000) for r in regrets]
+#   regrets = [max(r, -100) for r in regrets]
+#   print 'regrets', regrets
+#   ymax = max(ymax, max(regrets))
+#   ymin = min(ymin, min(regrets))
+#   plot(regrets)
+#   # plot([0, len(infLps) - 1], [origLp, origLp])
+# axis(ymin=ymin, ymax=ymax)
 
-axis(ymin=ymin, ymax=ymax)
+niters = len(run_results[0]['infLps'])
+
+def mean(xs):
+  return sum(xs) / len(xs)
+
+def quantile(xs, q):
+  xs = sorted(xs)
+  avg_index = (len(xs) - 1) * q
+  lo = xs[int(avg_index)]
+  hi = xs[1+int(avg_index)]
+  portion = avg_index - int(avg_index)
+  return hi * portion + lo * (1 - portion)
+
+burn = 1
+
+regret_distrs = [[r['origLp'] - r['infLps'][i] for r in run_results] for i in range(burn, niters)]
+print 'rd', len(regret_distrs), niters
+
+for q in [0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9]:
+  quants = []
+  for i in range(burn, niters):
[…]

### 2015-06-05 plots/plotdata.py

@@ -1,0 +1,124 @@
+from pylab import *
+import math
+from matplotlib.patches import Ellipse
+
+dat = eval(open('irisdata.txt').read())
+(xs, ys) = zip(*dat)
+
+# def iparams(p):
+#     (_, ((n1, n2), ((n3,),))) = p
+#     variance = -1 / (2 * n2)
+#     mean0 = n1 * variance
+#     mean1 = (n1 + n3) * variance
+#     return (variance, mean0, mean1)
+
+def boolcolor(x):
+    return 'red' if x else 'blue'
+
+params1 = [[{"stdev":0.5372731549542051,"baseMean":6.317781274586794,"coeffs":[-1.3016404423965577]},{"stdev":0.6880720208986841,"baseMean":4.984290584841217,"coeffs":[-3.3624798852527316]}]]
+
+params2 = [[{"stdev":0.6595119407561928,"baseMean":6.262000000000001,"coeffs":[]},{"stdev":0.8214401986754728,"baseMean":4.905999999999999,"coeffs":[]}],[{"stdev":0.3489469873777391,"baseMean":5.005999999999999,"coeffs":[]},{"stdev":0.17176728442867112,"baseMean":1.4640000000000004,"coeffs":[]}]]
+
+params3 = [{"stdev":0.6595119407561928,"baseMean":6.262000000000001,"coeffs":[]},{"stdev":0.3489469873777391,"baseMean":5.005999999999999,"coeffs":[]},{"stdev":0.460023458046562,"baseMean":-1.5557132215677991,"coeffs":[1.0318928811190988]},{"stdev":0.16567935943532425,"baseMean":0.8137676160441453,"coeffs":[0.12989060806149724]}]
+
+params3 = [{"stdev":0.7287275058887802,"baseMean":5.368444981174984,"coeffs":[]},{"stdev":0.6619425046955395,"baseMean":6.2631664887270455,"coeffs":[]},{"stdev":0.702275237872738,"baseMean":-6.875651795265539,"coeffs":[1.7159693631199096]},{"stdev":0.41252571793466597,"baseMean":-1.4608529214942643,"coeffs":[1.0341202813737411]}]
+
[…]

### 2015-06-05 src/expfam.js

@@ -87,6 +87,9 @@ var Double = {
       });
     });
   }),
+  randomDefault: fromMonad(function() {
+    return mcurry(global.sample, erp.gaussianERP, [0, 1]);
+  }),
   featuresMask: [true, false],
   mle: function(samples, natParam) {
     return gaussianMle(samples);
@@ -126,6 +129,9 @@ function Categorical(n) {
           return mreturn({base: base, weights: opt.transpose(weights)});
         });
       });
+    }),
+    randomDefault: fromMonad(function() {
+      return mcurry(global.sample, erp.randomIntegerERP, [n]);
     }),
     featuresMask: _.times(n-1, function() { return true; }),
     formatParams: function(params) { return params; }
@@ -190,6 +196,9 @@ function Tuple(types) {
         weights: util.concat(_.pluck(subParams, 'weights'))
       };
     },
+    randomDefault: mcurry(util.mapM, types, fromMonad(function(t) {
+      return t.randomDefault;
+    })),
     formatParams: function(params) {
       return types.map(function(type, i) {
         var subParams = {

### 2015-06-05 src/param_inference.js

@@ -14,6 +14,7 @@ var argv = require('yargs').argv;
 console.log('argv', argv);
 
 function DataSampler() {
+  this.index = 0;
   this.valueMap = {};
 }
 
@@ -22,7 +23,8 @@ DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var args = [].slice.call(arguments, 2, arguments.length);
   if (typeof addr == 'function') {
     return mbind(util.getAddress, function(a) {
-      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+      // return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', self.index++, addr, rf].concat(args));
     });
   }
 
@@ -48,6 +50,7 @@ var sampleData = fromMonad(function(s) {
 });
 
 function DataConditioner(valueMap) {
+  this.index = 0;
   this.valueMap = valueMap;
 }
 
@@ -56,7 +59,8 @@ DataConditioner.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var args = [].slice.call(arguments, 2, arguments.length);
   if (typeof addr == 'function') {
     return mbind(util.getAddress, function(a) {
-      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+      // return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', self.index++, addr, rf].concat(args));
     });
   }
 
@@ -116,9 +120,11 @@ function trainParameters(signature, calls, params) {
   return expfam.mle(retType, samps, params);
 }
 
-function UnknownParametersModel() {
+function UnknownParametersModel(opts) {
+  opts = opts || {};
   this.signatures = [];
   this.sampler = null;
+  this.logPartitionParticles = opts.logPartitionParticles || 30;
 }
 
 UnknownParametersModel.prototype.randFunction = function(s0, k0, a0) {
@@ -180,7 +186,7 @@ UnknownParametersModel.prototype.randSample = function(s, k, a, params) {
 
 UnknownParametersModel.prototype.logPartition = fromMonad(function(params) {
   var self = this;
-  var nparticles = 30;
[…]

### 2015-06-05 src/quipp.js

@@ -24,4 +24,5 @@ module.exports = {
   Tuple: util.makeWpplFunction(expfam.Tuple),
   Vector: Vector,
   Categorical: util.makeWpplFunction(expfam.Categorical),
+  randomValue: fromMonad(function(t) { return t.randomDefault; })
 };

### 2015-06-05 src/util.js

@@ -27,8 +27,11 @@ function withStackMessage(msg, x) {
 }
 
 function catchStack(x) {
+  assert(typeof x == 'function', x);
   try {
-    return x();
+    var result = x();
+    assert(typeof result == 'function', result);
+    return result;
   } catch (ex) {
     console.log(ex.stack);
     console.log(labelStack);
@@ -42,6 +45,7 @@ function getAddress(s, k, a) {
 }
 
 function trampoline(f) {
+  assert(typeof f == 'function', f);
   return function(s, k, a) {
     return function() {
       return f(s, k, a)();

### 2015-06-05 paper/paper.tex
stuff

@@
   \end{verbatim}
-  This model is estimated to assign probability density $e^{-393}$ to the data, or 0.0728 per point.
+  This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.
 
@@
   \end{center}
-  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, so it assigns significantly higher likelihood to the data.
+  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.
 
@@
 
-  This model is estimated to assign probability $e^{-269}$ to the data, or 0.166 per point, a large improvement from the previous model.
+  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.
 
@@
 
-\subsection{A more complex model}
-
-TODO: if there is time, we should put a more complex data science type example, where we add dependencies and show change in accuracy as we add/remove assumptions.
-
   \section{Discussion}

### 2015-06-05 paper/paper.tex
cleaned up examples

@@
 
-var pointType = Vector(dim, Double);
-var getPoint = randFunction(Categorical(nclusters), pointType);
+var Point = Vector(3, Double);
+var Cluster = Categorical(3);
+var getPoint = randFunction(Cluster, Point);
 return function() {
   return repeat(n, function(i) {
-    var cluster = randomInteger(nclusters);
-    return observe('point' + i, getPoint, cluster);
+    var cluster = randomValue(Cluster);
+    observe(getPoint, cluster);
   });
@@
 \begin{verbatim}
-var nfeatures = 10;
-var featuresType = Vector(nfeatures, Bool);
-var nclasses = 2;
 var n = 50;
 
-var class0Features = randFunction(featuresType);
-var class1Features = randFunction(featuresType);
+var Class = Categorical(2);
+var Features = Vector(10, Bool);
 
-var getFeatures = randFunction(Categorical(nclusters), pointType);
+var classFeatures = [randFunction(Features), randFunction(Features)];
+
 return function() {
   repeat(n, function(i) {
-    var whichClass = randomInteger(2);
-    if (whichClass == 0) {
-      observe('features' + i, class0Features);
-    } else {
-      observe('features' + i, class1Features);
-    }
+    var whichClass = randomValue(Class);
+    observe(classFeatures, whichClass);
   });
@@
 
-var factorType = Vector(nfactors, Double);
-var pointType = Vector(dim, Double);
-var getPoint = randFunction(factorType, pointType);
-var getFactors = function() {
-  return repeat(nfactors, function() { return gaussian(0, 1); });
-};
+var Factors = Vector(2, Double);
+var Point = Vector(5, Double);
+var getPoint = randFunction(Factors, Point);
+
 return function() {
   return repeat(n, function(i) {
-    var factors = getFactors();
-    return observe('point' + i, getPoint, factors);
+    var factors = randomValue(Factors);
+    observe(getPoint, factors);
   });
@@
 
-var nstates = 2;
-var nobs = 3;
+var chainLength = 20;
 
-var chainLength = 8;
+var n = 10;
 
-var n = 30;
+var State = Categorical(2);
+var Obs = Categorical(4);
+var transFun = randFunction(State, State);
+var obsFun = randFunction(State, Obs);
 
-var stateType = Categorical(nstates);
-var obsType = Categorical(nobs);
-var transFun = randFunction(stateType, stateType);
-var obsFun = randFunction(stateType, obsType);
-
-var observeStates = function(sampIndex, startState, i) {
+var observeStates = function(startState, i) {
   if (i == chainLength) {
@@
   } else {
-    observe('obs' + sampIndex + '_' + i, obsFun, startState);
-    return [startState].concat(observeStates(sampIndex, transFun(startState), i+1));
+    observe(obsFun, startState);
+    observeStates(transFun(startState), i+1);
   }
@@
   return repeat(n, function(sampIndex) {
-    return observeStates(sampIndex, randomInteger(nstates), 0);
+    observeStates(randomValue(State), 0);
   });
@@
 \begin{verbatim}
-var nClasses = 2;
-var nWords = 100;
-var maxWordsPerDocument = 1000;
+
+var maxWordsPerDocument = 100;
 var nDocuments = 20;
 
-var classType = Categorical(nClasses);
-var wordType = Categorical(nWords);
+var Class = Categorical(3);
+var Word = Categorical(10);
+var classToWord = randFunction(Class, Word);
 
-var classToWord = randFunction(ClassType, WordType);
 return function() {
-  return repeat(nDocuments, function(docIndex) {
-    var whichClass = randomInteger(nClasses);
-    var nWords = observe('len' + docIndex, randomInteger, maxWordsPerDocument);
-    repeat(nWordsPerDocument, function(wordIndex) {
-      observe('word' + doc + '_' + wordIndex, classToWord, whichClass);
+  return repeat(nDocuments, function() {
+    var whichClass = randomValue(Class);
+    var nWords = observe(randomIntegerERP, maxWordsPerDocument);
+    repeat(nWords, function(wordIndex) {
+      observe(classToWord, whichClass);
     });
@@
 };
+
 \end{verbatim}
@@
 \begin{verbatim}
-var inputDim = 100;
-var hiddenDim = 20;
-var outputDim = 2;
 
-var inputType = Categorical(inputDim);
-var hiddenType = Categorical(hiddenDim);
-var outputType = Categorical(outputDim);
+var Input = Categorical(30);
+var Hidden = Categorical(10);
+var Output = Categorical(2);
 
@@
 
-testParamInference(function(randFunction) {
-  var inputToHidden = randFunction(inputType, hiddenType);
-  var hiddenToOutput = randFunction(hiddenType, outputType);
+var inputToHidden = randFunction(Input, Hidden);
+var hiddenToOutput = randFunction(Hidden, Output);
 
@@
     var hiddenLayer = inputToHidden(inputs[sampIndex]);
-      observe('output' + sampIndex, hiddenToOutput, hiddenLayer);
+    observe(hiddenToOutput, hiddenLayer);
   });
 };
-});
 \end{verbatim}

### 2015-06-05 paper/paper.tex
example cleanup: randFunction(Cluster, Point)

@@
 var Point = Vector(2, Double);
-var getPoint = randFunction(Categorical(2), pointType);
+var getPoint = randFunction(Cluster, Point);
 return function() {

### 2015-06-05 paper/paper.tex
charts and stuff

--- 2015-06-05 paper/paper.tex
@@
 \usepackage{amsmath}
+% \usepackage{minted}
 \usepackage{hyperref}
@@
   \end{abstract}
+
+   % TODO syntax coloring
 
@@
   sample from the given cluster.  The \texttt{observe} function allows us to
-  observe data; here it says that the observation named \texttt{point3} is equal
-  to the result of calling \texttt{getPoint(cluster)}, where \texttt{cluster} is
-  cluster number 3.
+  observe data; here it says that our observations consist of \texttt{getPoint(cluster)} for
+  each cluster.
 
@@
   var cluster = randomValue(Cluster);
-  observe('point' + i, getPointFunctions[cluster]);
+  observe(getPointFunctions[cluster]);
 });
@@
 
-  Note that the clusters are still restricted to have $x$ and $y$ be independent.  To allow these coordinates
-  to be correlated within a cluster, we can use the following model:
+  Note that the clusters represent axis-aligned multivariate Gaussian distributions.   To allow them to be arbitrary Gaussian distributions (where the $x$ and $y$ coordinates may be correlated),
+  we can use the following model:
 
@@
 
-    For unknown functions, the family of random functions used is a kind
-    of generalized linear model.  We assume that the distribution
+    \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
+    determined by the desired argument types and return type.
+    The distribution
     of the function's return value $y$
@@
 
-    To infer both latent variables and parameters, we run the
-    expectation maximization algorithm on the probabilistic model, iterating stages of
+    To infer both latent variables and parameters, we use a Monte Carlo
+    expectation maximization algorithm \cite{mcem} on the probabilistic model, iterating stages of
     estimating latent variables using Metropolis Hastings and inferring
@@
 
-We randomly generated parameters for this example 100 times, and each time took 100 samples and then ran 10 EM iterations.
+\begin{center}
+  \includegraphics[scale=0.5]{../plots/accuracy_nd_clustering.png}
+\end{center}
+
+% We randomly generated parameters for this example 100 times, and each time took 100 samples and then ran 10 EM iterations.
 The accuracy is defined as the maximum percentage of points assigned to the correct cluster, for any permutation of clusters.
 On average, accuracy increased in each EM iteration, as shown it this graph:
-
-\begin{center}
-\includegraphics[scale=0.4]{cluster_accuracy.png}
-\end{center}
+% 
+% \begin{center}
+% \includegraphics[scale=0.4]{cluster_accuracy.png}
+% \end{center}
 
@@
 
+\begin{center}
+  \includegraphics[scale=0.5]{../plots/accuracy_factor_analysis.png}
+\end{center}
+
   \subsection{Hidden Markov model}
@@
 
+\begin{center}
+  \includegraphics[scale=0.5]{../plots/accuracy_hmm.png}
+\end{center}
+
 \subsection{Latent Dirichlet allocation}
@@
 
-(figure should show accuracy over time.  Accuracy can be measured as distance between the learned categorical distributions, for some permutation of classes)
-
+\begin{center}
+  \includegraphics[scale=0.5]{../plots/accuracy_lda.png}
+\end{center}
 \subsection{Neural network}

### 2015-06-05 paper/paper.tex
syntax highlight

@@
 \usepackage{cite}
+\usepackage{listings}
 \usepackage{amsmath}
-% \usepackage{minted}
+% \usepackage{minted1}
 \usepackage{hyperref}
@@
   Here is an example of a Quipp program to cluster 2d points into 2 clusters:
-  \begin{verbatim}
+\lstdefinelanguage{JavaScript}{
+      morekeywords={break, case, catch, continue, debugger, default, delete,         do, else, false, finally, for, function, if, in, instanceof, new, null, return, switch, this, throw, true, try, typeof, var, void, while, with},
+      morecomment=[s]{/*}{*/},
+      morecomment=[l]//,
+      morestring=[b]",
+      morestring=[b]'
+    }
+  \lstset{language=JavaScript, xleftmargin=0.5in}
+
+  \begin{lstlisting}
 var n = 100;
@@
 };
-  \end{verbatim}
+  \end{lstlisting}
 
@@
   We could use these parameters to fill in the generative model:
-  \begin{verbatim}
+  \begin{lstlisting}
 return repeat(n, function(i) {
@@
 });
-  \end{verbatim}
+  \end{lstlisting}
   This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.
@@
   model, which uses a separate random function for each cluster:
-  \begin{verbatim}
+  \begin{lstlisting}
 var Cluster = Categorical(2);
@@
 });
-\end{verbatim}
+\end{lstlisting}
   Using this model, we get the following clusters:
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 
@@
 };
-\end{verbatim}
+\end{lstlisting}
 
@@
 {\small
-\begin{verbatim}
+\begin{lstlisting}
 var nclusters = 3;
@@
 };
-\end{verbatim}
+\end{lstlisting}
 }
@@
 {\small
-\begin{verbatim}
+\begin{lstlisting}
 var n = 50;
@@
 {\small
-\begin{verbatim}
+\begin{lstlisting}
 
@@
 
-\end{verbatim}
+\end{lstlisting}
 }

### 2015-06-05 paper/paper.tex
cleaned up models

@@
 \begin{lstlisting}
-var nclusters = 3;
-var dim = 2;
+var Cluster = Categorical(3);
+var Point = Vector(2, Double);
+var getPoint = randFunction(Cluster, Point);
 
-var n = 100;
-
-var Point = Vector(3, Double);
-var Cluster = Categorical(3);
-var getPoint = randFunction(Cluster, Point);
-return function() {
-  return repeat(n, function(i) {
+var model = function() {
   var cluster = randomValue(Cluster);
   observe(getPoint, cluster);
-  });
 };
@@
 \begin{lstlisting}
-var n = 50;
-
 var Class = Categorical(2);
 var Features = Vector(10, Bool);
-
 var classFeatures = [randFunction(Features), randFunction(Features)];
 
-return function() {
-  repeat(n, function(i) {
+var model = function() {
   var whichClass = randomValue(Class);
   observe(classFeatures, whichClass);
-  });
 };
@@
 \begin{lstlisting}
-
-var nfactors = 2;
-var dim = 4;
-
-var n = 50;
-
 var Factors = Vector(2, Double);
@@
 
-return function() {
-  return repeat(n, function(i) {
+var model = function() {
   var factors = randomValue(Factors);
-    observe(getPoint, factors);
-  });
+  return observe(getPoint, factors);
 };
@@
 \begin{lstlisting}
-
 var chainLength = 20;
-
-var n = 10;
-
 var State = Categorical(2);
@@
 
-return function() {
-  return repeat(n, function(sampIndex) {
-    observeStates(randomValue(State), 0);
-  });
+var model = function() {
+  return observeStates(randomValue(State), 0);
 };
@@
 \begin{lstlisting}
-
 var maxWordsPerDocument = 100;
-var nDocuments = 20;
-
 var Class = Categorical(3);
@@
 
-return function() {
-  return repeat(nDocuments, function() {
+var model = function() {
   var whichClass = randomValue(Class);
-    var nWords = observe(randomIntegerERP, maxWordsPerDocument);
+  var nWords = observe('len' + docIndex, randomIntegerERP, maxWordsPerDocument);
   repeat(nWords, function(wordIndex) {
@@
   });
-    return whichClass;
-  });
 };
-
 \end{lstlisting}
@@
 \begin{lstlisting}
-
-var Input = Categorical(30);
-var Hidden = Categorical(10);
-var Output = Categorical(2);
-
-var inputs = [...];
+var Input = Vector(30, Bool);
+var Hidden = Vector(10, Bool);
+var Output = Bool;
 
@@
 
-return function() {
-  return repeat(inputs.length, function(sampIndex) {
+var model = function() {
+  var inputLayer = randomValue(Input);
   var hiddenLayer = inputToHidden(inputs[sampIndex]);
   observe(hiddenToOutput, hiddenLayer);
-  });
 };

### 2015-06-05 src/param_inference.js

@@ -18,13 +18,24 @@ function DataSampler() {
   this.valueMap = {};
 }
 
+var nextObsIndex = function(s, k, a) {
+  var res = s._obsIndex++;
+  return k(s, res);
+};
+
+var resetObsIndex = function(s, k, a) {
+  s = _.clone(s);
+  s._obsIndex = 0;
+  return k(s);
+};
+
 DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var self = this;
   var args = [].slice.call(arguments, 2, arguments.length);
   if (typeof addr == 'function') {
-    return mbind(util.getAddress, function(a) {
-      // return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
-      return mcurryMethod.apply(null, [self, 'sampleOrCondition', self.index++, addr, rf].concat(args));
+    // return mbind(util.getAddress, function(a) {
+    return mbind(nextObsIndex, function(a) {
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
     });
   }
 
@@ -42,10 +53,12 @@ DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
 });
 
 var sampleData = fromMonad(function(s) {
-  var sampler = new DataSampler();
-  global.observe = sampler.sampleOrCondition.bind(sampler);
-  return mbind(s, function(res) {
-    return mreturn(sampler.valueMap);
+  return mbind(resetObsIndex, function() {
+    var sampler = new DataSampler();
+    global.observe = sampler.sampleOrCondition.bind(sampler);
+    return mbind(s, function(res) {
+      return mreturn(sampler.valueMap);
+    });
   });
 });
 
@@ -58,9 +71,9 @@ DataConditioner.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var self = this;
   var args = [].slice.call(arguments, 2, arguments.length);
   if (typeof addr == 'function') {
-    return mbind(util.getAddress, function(a) {
-      // return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
-      return mcurryMethod.apply(null, [self, 'sampleOrCondition', self.index++, addr, rf].concat(args));
+    return mbind(nextObsIndex, function(a) {
+    // return mbind(util.getAddress, function(a) {
[…]

### 2015-06-05 src/quipp.js

@@ -3,7 +3,7 @@ var param_inference = require('./param_inference');
 var expfam = require('./expfam');
 var util = require('./util');
 
-var fromMonad = util.fromMonad, mreturn = util.mreturn;
+var fromMonad = util.fromMonad, mreturn = util.mreturn, mcurry = util.mcurry;
 
 function factorScore(s, k, a, fn) {
   var args = [].slice.call(arguments, 4);
@@ -14,6 +14,8 @@ var Vector = fromMonad(function(n, t) {
   return mreturn(expfam.Tuple(_.times(n, function() { return t; })));
 });
 
+var Bool = mcurry(util.Categorical, 2);
+
 
 module.exports = {
   factorScore: factorScore,
@@ -21,6 +23,7 @@ module.exports = {
   testParamInference: param_inference.testParamInference,
   testParamInferenceSplit: param_inference.testParamInferenceSplit,
   Double: expfam.Double,
+  Bool: Bool,
   Tuple: util.makeWpplFunction(expfam.Tuple),
   Vector: Vector,
   Categorical: util.makeWpplFunction(expfam.Categorical),

### 2015-06-05 paper/paper.tex
more model cleanup

@@
 \DeclareMathOperator*{\Prob}{\mathbf{P}}
+
+\lstdefinelanguage{JavaScript}{
+      morekeywords={break, case, catch, continue, debugger, default, delete,         do, else, false, finally, for, function, if, in, instanceof, new, null, return, switch, this, throw, true, try, typeof, var, void, while, with},
+      morecomment=[s]{/*}{*/},
+      morecomment=[l]//,
+      morestring=[b]",
+      morestring=[b]'
+    }
+  \lstset{language=JavaScript, xleftmargin=0.5in}
 
@@
   Here is an example of a Quipp program to cluster 2d points into 2 clusters:
-\lstdefinelanguage{JavaScript}{
-      morekeywords={break, case, catch, continue, debugger, default, delete,         do, else, false, finally, for, function, if, in, instanceof, new, null, return, switch, this, throw, true, try, typeof, var, void, while, with},
-      morecomment=[s]{/*}{*/},
-      morecomment=[l]//,
-      morestring=[b]",
-      morestring=[b]'
-    }
-  \lstset{language=JavaScript, xleftmargin=0.5in}
 
   \begin{lstlisting}
-var n = 100;
-
 var Cluster = Categorical(2);
@@
 var getPoint = randFunction(Cluster, Point);
-return function() {
-  repeat(n, function(i) {
+
+var model = function() {
   var cluster = randomValue(Cluster);
   observe(getPoint, cluster);
-  });
 };
@@
   \begin{lstlisting}
-return repeat(n, function(i) {
+var model = function() {
   var cluster = randomInteger(nclusters);
@@
           gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
-});
+};
   \end{lstlisting}
@@
 var getPointFunctions = [randFunction(Point), randFunction(Point)];
-return repeat(n, function(i) {
+
+var model = function() {
   var cluster = randomValue(Cluster);
   observe(getPointFunctions[cluster]);
-});
+};
 \end{lstlisting}
@@
 \begin{lstlisting}
-
 var Cluster = Categorical(2);
@@
 
-return function() {
-  return repeat(n, function(i) {
+var model = function() {
   var cluster = randomInteger(nclusters);
@@
   var y = observe(getY[cluster], x);
-  });
 };

### 2015-06-05 paper/paper.tex
cleanup

@@
 
-  It is written as a generative model for the observations.  Notice that
-  we declared two types (\texttt{Point} and \texttt{Cluster}) and one
-  random function \texttt{getPoint}).  Type annotations are necessary for random
+  We declared two types (\texttt{Point} and \texttt{Cluster}) and one
+  random function (\texttt{getPoint}).  Type annotations are necessary for random
   functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
@@
   fact that there are 2 possible clusters (so a cluster may be either 0 or 1).
-  We use the \texttt{randomValue} function to generate a random uniform cluster. We do not know
-  the distribution of points in each cluster, represented by the fact that \texttt{getPoint}
-  is an unknown random function.  We will fill in the
-  \texttt{getPoint} function with a learned function that will take a random
-  sample from the given cluster.  The \texttt{observe} function allows us to
-  observe data; here it says that our observations consist of \texttt{getPoint(cluster)} for
-  each cluster.
+
+  The variable \texttt{model} specifies a generative model for a single data point.
+  We use the \texttt{randomValue} function to generate a random uniform cluster, and then
+  use \texttt{getPoint} to generate the point given the cluster.  Since \texttt{getPoint}
+  is an unknown random function, we will need to infer its parameters.
+  The \texttt{observe} function allows us to observe data; here it says that our
+  observation consist of the result of the call \texttt{getPoint(cluster)}.
 
@@
 
-  Note that, because the two clusters are forced to have the same variance, they do not fit the data well,
-  since the data has a different shape in each location.  To fix this problem, we can substitute the following
-  model, which uses a separate random function for each cluster:
+  \section{Family of distributions}
+
+  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned Gaussian clusters with equal variances.
+In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
+    determined by the desired argument types and return type.
+    The distribution
+    of the function's return value $y$
+    is some exponential family whose natural
+    parameters are determined from the arguments $x$:
+
+    $$p_{\eta}(y | x) = \exp\left(\eta(x)^T \phi(y) - g(\eta(x))\right)$$
+
+    Here, $\eta(x)$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
+    and $g$ is the log partition function.
+
+    To determine $\eta(x)$, we label
+    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
+    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
+    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
+    and features are both $[X = 1], [X = 2], ..., [X=n - 1]$.
+    The natural
+    parameters corresponding to non-features are constant, while natural
+    parameters corresponding to features are determined as an affine
+    function of the features of the arguments.
+
+    Let $\psi(x)$ be the features of $x$.  Then
+  $$\eta(x) = \mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}$$
+    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$
+
+    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
+    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
+    This ensures that only the
+    natural parameters that are features of $y$ are affected by $x$.  The following shows the types
+    in Quipp and their corresponding sufficient statistics, features, and English descriptions.
+
+    \begin{tabular}{| c | c | c | c|}
+      \hline
+      T & $\phi_T(X)$ & $\psi_T(X)$ & Random function class \\
+
+      \hline
+    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & Linear regression with Gaussian noise \\
+
+      \hline
+    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $n$-class logistic regression \\
+
+      \hline
+    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
+                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
+      Independent regression of each component
+      \\
+      \hline
+    \end{tabular}
+
+    Note that $\texttt{Vector(n, T)}$ is shorthand for $Tuple(T, T, ..., T)$, with $n$ copies of $T$.
+
+  \section{Example: improving the clustering model}
+  Using this knowledge about the form of the random functions, we can improve the clustering model from before.
+  As observed in the previous graph, the two clusters are forced to have the
+  same variance. They do not fit the data well, since the data has a different
+  shape in each location.  To fix this problem, we can substitute the following
+  model, which uses a separate random function (and therefore a separate axis-aligned Gaussian distribution) for each cluster:
   \begin{lstlisting}
@@
 
-  Note that the clusters represent axis-aligned multivariate Gaussian distributions.   To allow them to be arbitrary Gaussian distributions (where the $x$ and $y$ coordinates may be correlated),
+  The fact that the gaussian distribution for each cluster must be axis-aligned limits the degree to which the model can fit the data.
+To allow each cluster to be an arbitrary Gaussian distributions (where the $x$ and $y$ coordinates may be correlated),
   we can use the following model:
@@
 
-  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend on the $x$ coordinate).
+  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).
 This allows $x$ and $y$ to be correlated, improving fit:
@@
   This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.
-
-  \section{Family of distributions}
-
-    \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
-    determined by the desired argument types and return type.
-    The distribution
-    of the function's return value $y$
-    is some exponential family whose natural
-    parameters are determined from the arguments $x$:
-
-    $$p_{\eta}(y | x) = \exp\left(\eta(x)^T \phi(y) - g(\eta(x))\right)$$
-
-    Here, $\eta(x)$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
-    and $g$ is the log partition function.
-
-    To determine $\eta(x)$, we label
-    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
-    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
-    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
-    and features are both $[X = 1], [X = 2] ..., [X=n - 1]$.
-    The natural
-    parameters corresponding to non-features are constant, while natural
-    parameters corresponding to features are determined as an affine
-    function of the features of the arguments.
-
-    Let $\psi(x)$ be the features of $x$.  Then
-  $$\eta(x) = \mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}$$
-    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$
-
-    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
-    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
-    This ensures that only the
-    natural parameters that are features of $y$ are affected by $x$.
-
-    \begin{tabular}{| c | c | c | c|}
-      \hline
-      T & $\phi_T(X)$ & $\psi_T(X)$ & Random function class \\
-
-      \hline
-    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & Linear regression with Gaussian noise \\
-
-      \hline
-    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $n$-class logistic regression \\
-
-      \hline
-    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
-                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
-      Independent prediction of each component
-      \\
-      \hline
-    \end{tabular}
 
@@
     For the expectation step, we must estimate latent variable distributions given
-    fixed values for the parameters.  To do this, we can run the Quipp program
-    with random functions set to use these fixed parameter values to generate their results.
+    fixed values for the parameters.  To do this, we can replace unknown random functions in the Quipp program
+    with random functions set to use these fixed parameter values, yielding a probabilistic program.
     We use the Metropolis Hastings algorithm to perform inference in this program,
-    yielding traces.  Next, for each random function, we can find all calls
+    yielding a distribution of execution traces (where each execution trace specifies the result of every
+    call to a random function).  Next, for each random function, we can find all calls
     to it in the trace to get the training data.
@@
     Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
-    problem because the log probability function is concave:
+    problem because the log probability function is concave as a function of $\mathbf{N}$:
     $$\log p_{\mathbf{N}}(y | x) = \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)$$

### 2015-06-05 paper/paper.tex
fix tables

@@
 \usepackage{amsmath}
+\usepackage{array}
 % \usepackage{minted1}
@@
 \setlength{\emergencystretch}{10em}
+
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 
-  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned Gaussian clusters with equal variances.
+  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned 2-dimensional Gaussian clusters with equal variances.
 In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
@@
     This ensures that only the
-    natural parameters that are features of $y$ are affected by $x$.  The following shows the types
+    natural parameters that are features of $y$ are affected by $x$.
+
+    Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
+    returns a sample from the ``default'' distribution for a given type.  The following shows the types
     in Quipp and their corresponding sufficient statistics, features, and English descriptions.
 
-    \begin{tabular}{| c | c | c | c|}
+    \begin{center}
+    {\scriptsize
+      \begin{tabular}{| c | c | c | c | C{3cm} |}
       \hline
-      T & $\phi_T(X)$ & $\psi_T(X)$ & Random function class \\
+      T & $\phi_T(X)$ & $\psi_T(X)$ & \texttt{randomValue(T)} & Random function class \\
 
       \hline
-    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & Linear regression with Gaussian noise \\
+    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & 
+      $\mathcal{N}(0, 1)$ &
+      Linear regression\\
 
       \hline
-    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $n$-class logistic regression \\
+    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & uniform & $n$-class logistic regression \\
 
@@
                            & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
-      Independent regression of each component
+      \begin{tabular}{l}
+        \texttt{[randomValue(T1),}
+       \\
+     \texttt{ randomValue(T2)]}
+      \end{tabular}
+                           &
+      Independent regression
       \\
@@
     \end{tabular}
+  }
+\end{center}
 
-    Note that $\texttt{Vector(n, T)}$ is shorthand for $Tuple(T, T, ..., T)$, with $n$ copies of $T$.
+  Note that \texttt{Vector(n, T)} is shorthand for \texttt{Tuple([T, T, ..., T])} (with $n$ copies of $T$),
+  and \texttt{Bool} is shorthand for \texttt{Categorical(2)}.
 
@@
     estimating latent variables using Metropolis Hastings and inferring
-    parameters using gradient descent.
+    parameters using gradient descent.  The first iteration uses randomly generated parameters.
 
@@
       \item
-        Estimate $\log P(x_{test} | \hat{\theta})$ and compare to $\log P(x_{test} | \theta)$
+        Estimate the regret $\log P(x_{test} | \hat{\theta}) - \log P(x_{test} | \theta)$
     \end{itemize}
     Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
-    We use the Sequential Monte Carlo algorithm for this.  Between observations,
+    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.
 

### 2015-06-05 src/quipp.js

@@ -14,7 +14,11 @@ var Vector = fromMonad(function(n, t) {
   return mreturn(expfam.Tuple(_.times(n, function() { return t; })));
 });
 
-var Bool = mcurry(util.Categorical, 2);
+var Tuple2 = fromMonad(function(args) {
+  return mreturn(expfam.Tuple([].slice.call(args)));
+});
+
+var Bool = util.Categorical(2);
 
 
 module.exports = {
@@ -24,7 +28,7 @@ module.exports = {
   testParamInferenceSplit: param_inference.testParamInferenceSplit,
   Double: expfam.Double,
   Bool: Bool,
-  Tuple: util.makeWpplFunction(expfam.Tuple),
+  Tuple: Tuple2,
   Vector: Vector,
   Categorical: util.makeWpplFunction(expfam.Categorical),
   randomValue: fromMonad(function(t) { return t.randomDefault; })

### 2015-06-05 paper/paper.tex
paper paper

@@
   \begin{abstract}
-    In probabilistic programs, sometimes it is difficult to specify the correct
+    When writing probabilistic programs, sometimes it is difficult for the programmer to specify the correct
     parameterized family of distributions.  We explore an extension to
     probabilistic programming languages that allows programmers to mark some
-    distributions as unspecified.  Then, we can fill in the distribution with
-    some family and infer parameters.
+    distributions as unspecified.  Then, for each distribution, the system can
+    determine some reasonable family of distributions and infer maximum likelihood
+    parameters.
   \end{abstract}
-
-   % TODO syntax coloring
 
@@
       \item
-        Estimate the regret $\log P(x_{test} | \hat{\theta}) - \log P(x_{test} | \theta)$
+        Estimate $\log P(x_{test} | \hat{\theta})$
+      \item
+        Use the previous estimates to estimate the regret  $\log P(x_{test} | \hat{\theta}) - \log P(x_{test} | \theta)$
     \end{itemize}

### 2015-06-05 plotaccuracy.py
running lda

@@ -58,7 +58,7 @@ for q in [0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9]:
   for i in range(burn, niters):
     regrets = regret_distrs[i-burn]
     quants.append(max(0, quantile(regrets, q)))
-  plot(quants)
+  plot(range(burn, niters), quants)
   
 
 

### 2015-06-05 src/expfam.js

@@ -101,6 +101,15 @@ var Double = {
   }
 };
 
+var randCategoricalNp = fromMonad(function(n) {
+  var alphas = _.times(n, function() { return 1; });
+  return mbind(global.sample, erp.dirichletERP, alphas, function(p) {
+    return mreturn(_.range(1, n).map(function(i) {
+      return Math.log(p[i] / p[0]);
+    }));
+  });
+});
+
 function Categorical(n) {
   assert(typeof n == 'number');
   return {
@@ -119,11 +128,9 @@ function Categorical(n) {
       return global.sample(s, k, a, erp.discreteERP, [probs]);
     },
     defaultNatParam: _.times(n-1, function() { return 0.0; }),
-    randNatParam: fromMonad(function() {
-      return mcurry(util.replicateM, n-1, mcurry(global.sample, erp.gaussianERP, [0, 1]));
-    }),
+    randNatParam: mcurry(randCategoricalNp, n),
     randParams: fromMonad(function(nfeatures) {
-      var getRow = mcurry(util.replicateM, n-1, mcurry(global.sample, erp.gaussianERP, [0, 1]));
+      var getRow = mcurry(randCategoricalNp, n);
       return mbind(getRow, function(base) {
         return mbind(util.replicateM, nfeatures, getRow, function(weights) {
           return mreturn({base: base, weights: opt.transpose(weights)});

### 2015-06-05 src/param_inference.js

@@ -32,7 +32,7 @@ var resetObsIndex = function(s, k, a) {
 DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var self = this;
   var args = [].slice.call(arguments, 2, arguments.length);
-  if (typeof addr == 'function') {
+  if (typeof addr == 'function' || typeof addr == 'object') {
     // return mbind(util.getAddress, function(a) {
     return mbind(nextObsIndex, function(a) {
       return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
@@ -70,7 +70,7 @@ function DataConditioner(valueMap) {
 DataConditioner.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var self = this;
   var args = [].slice.call(arguments, 2, arguments.length);
-  if (typeof addr == 'function') {
+  if (typeof addr == 'function' || typeof addr == 'object') {
     return mbind(nextObsIndex, function(a) {
     // return mbind(util.getAddress, function(a) {
       return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
@@ -318,9 +318,9 @@ var testParamInferenceSplit = fromMonad(function(fun, opts) {
       console.log('orig params', JSON.stringify(upmWrong.formatParameters(origParams)));
       console.log('also', JSON.stringify(origParams));
       return mbind(generateRandData, fun, opts, origParams, function(trainingData) {
-        console.log('training data', trainingData);
+        // console.log('training data', trainingData);
         return mbind(generateRandData, fun, opts, origParams, function(testData) {
-          console.log('test data', testData);
+          // console.log('test data', testData);
 
           function innerSamplerForData(data) {
             return fromMonad(function(randFunction) {

### 2015-06-05 src/quipp.js

@@ -18,7 +18,7 @@ var Tuple2 = fromMonad(function(args) {
   return mreturn(expfam.Tuple([].slice.call(args)));
 });
 
-var Bool = util.Categorical(2);
+var Bool = expfam.Categorical(2);
 
 
 module.exports = {

### 2015-06-05 paper/paper.tex
changes changes

@@
   functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
-  points are 2-dimensional, and the type \texttt{Categorical(2)} expresses the
+  points are vectors in $\mathbb{R}^2$, and the type \texttt{Categorical(2)} expresses the
   fact that there are 2 possible clusters (so a cluster may be either 0 or 1).
@@
   have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
-  We could use these parameters to fill in the generative model:
+  We could use these parameters to fill in the generative model, giving us a webppl program:
   \begin{lstlisting}
@@
     function of the features of the arguments.
+    Features are selected so they correspond
+    to natural parameters that can take on any real number.
 
@@
     where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
-    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
+    to a sufficient statistic of $y$ that is not a feature and whose column is not 1.
     This ensures that only the
@@
       \item
-        Use the EM algorithm to infer approximate parameters $\hat{\theta}$ from $x_{train}$
+        Use the EM algorithm to infer approximate parameters $\hat{\theta}_i$ from $x_{train}$, for iteration $i = 1, 2, ...$
       \item
-        Estimate $\log P(x_{test} | \hat{\theta})$
+        Estimate $\log P(x_{test} | \hat{\theta}_i)$
       \item
-        Use the previous estimates to estimate the regret  $\log P(x_{test} | \hat{\theta}) - \log P(x_{test} | \theta)$
+        Use the previous estimates to estimate the regret $\log P(x_{test} | \hat{\theta}_i) - \log P(x_{test} | \theta)$
     \end{itemize}
@@
   \subsection{Clustering}
+
 {\small
@@
 The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.
-This results in the model attempting to find each point as an affine function of a vector of standard normal values.
+\texttt{randomValue(Factors)} will return a vector from the standard multivariate normal distribution.
+This results in the model attempting to predict each point as an affine function of a vector of standard normal values, with Gaussian noise.
 
@@
 In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.
-This means that we will learn both the state transitions and the observation distribution.
+This means that we will learn both the state transition matrix and the observation matrix.
 
@@
 var maxWordsPerDocument = 100;
-var Class = Categorical(3);
+var Topic = Categorical(3);
 var Word = Categorical(10);
-var classToWord = randFunction(Class, Word);
+var topicToWord = randFunction(Topic, Word);
 
 var model = function() {
-  var whichClass = randomValue(Class);
-  var nWords = observe('len' + docIndex, randomIntegerERP, maxWordsPerDocument);
+  var whichClass = randomValue(Topic);
+  var nWords = observe(randomIntegerERP, maxWordsPerDocument);
   repeat(nWords, function(wordIndex) {
@@
 
-In this example, we use the unknown function \texttt{classToWord} to map classes to word distributions.
-Note that each column of the matrix of parameters for \texttt{classToWord} will represent a categorical distribution over words, and there will be one column for each class.
+For latent Dirichlet allocation, we use the unknown function \texttt{topicToWord} to map latent topics to word distributions.  We will learn a different categorical distribution for each topic.
 
@@
 
-  We have found that it is possible to write many useful machine learning models as Quipp programs and then use generic algorithms for inference.
-Furthermore, performance is <???>.
-This should make it much easier for non-experts to write useful machine learning models.
+  It is possible to write many useful machine learning models as Quipp programs and then use a generic Monte Carlo expectation maximization algorithm for inference.
+This algorithm infers parameters reasonably well.
+Non-experts will find it easier to write useful machine learning models in this language compared to other probabilistic programming lanvguages.
 
-  In the future, it will be useful to expand the set of types supported.
-It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.
-Also, it will be useful to create a more usable interface to infer parameters and perform additional data pracessing given these parameters.
+  In the future, it will be useful to support additional types.
+For example, we could add types for standard exponential families such as Poisson and Gamma.
+Supporting disjoint union types is also straightforward.
+Also, it will be useful to create a more usable interface to infer parameters and perform additional data processing given these parameters.
 

### 2015-06-05 getdata.py

@@ -13,6 +13,8 @@ if program_name == 'factor_analysis':
   niters = 12
 if program_name == 'hmm':
   niters = 10
+if program_name == 'lda':
+  niters = 10
 
 
 

### 2015-06-05 plotaccuracy.py

@@ -1,4 +1,5 @@
 from pylab import *
+from matplotlib import pyplot as plt
 import sys
 import sys
 import json
@@ -53,14 +54,21 @@ burn = 1
 regret_distrs = [[r['origLp'] - r['infLps'][i] for r in run_results] for i in range(burn, niters)]
 print 'rd', len(regret_distrs), niters
 
-for q in [0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9]:
+xlabel('iteration')
+ylabel('regret')
+
+quantiles = [0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9]
+quantiles = [0.1, 0.3, 0.5, 0.7, 0.9]
+for q in quantiles:
   quants = []
   for i in range(burn, niters):
     regrets = regret_distrs[i-burn]
     quants.append(max(0, quantile(regrets, q)))
-  plot(range(burn, niters), quants)
+  plot(range(burn, niters), quants, label=str(q))
   
 
+title('Regret quantiles by iteration')
+legend()
 
 
 plot_name = 'plots/accuracy_' + program_name + '.png'

### 2015-06-05 paper/paper.tex
paperrr

@@
 \usepackage{amsmath}
+\usepackage{placeins}
 \usepackage{array}
@@
 
-  In an ordinary probabilistic programming language (such as Church),
-  it is possible to treat parameters as random variables.  This
-  would allow ordinary inference algorithms to infer parameters.  However,
-  there are advantages of having unknown functions as a feature
-  in the language.
-  First, it is easier to
-  write programs without knowing the details of different parameterized distributions.
-  Second, the system can use specialized algorithms to infer parameters faster.
+  % In an ordinary probabilistic programming language (such as Church),
+  % it is possible to treat parameters as random variables.  This
+  % would allow ordinary inference algorithms to infer parameters.  However,
+  % there are advantages of having unknown functions as a feature
+  % in the language.
+  % First, it is easier to
+  % write programs without knowing the details of different parameterized distributions.
+  % Second, the system can use specialized algorithms to infer parameters faster.
 
@@
 
+  {
   \begin{lstlisting}
@@
   \end{lstlisting}
+}
 
@@
 
-  To demonstrate, let us run this example on a dataset consisting of 150 points (TODO cite).  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
+  To demonstrate, let us run this example on the Iris dataset, which consists of 150 points \cite{Lichman:2013}.
+When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
   In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
-  split the data into 2 clusters with equal variance:
+  split the data into 2 clusters with equal variance, as seen in Figure 1.
+
+  \begin{figure}[h]
   \begin{center}
@@
   \end{center}
+  \caption{Clustering the Iris dataset}
+\end{figure}
 
@@
   We could use these parameters to fill in the generative model, giving us a webppl program:
+  {
   \begin{lstlisting}
@@
   \end{lstlisting}
+}
   This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.
@@
 
-  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned 2-dimensional Gaussian clusters with equal variances.
+  In the previous example, \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 unknown axis-aligned 2-dimensional Gaussian clusters with equal variances.
 In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
@@
     Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
-    returns a sample from the ``default'' distribution for a given type.  The following shows the types
+    returns a sample from the ``default'' distribution for a given type.  Table 1 shows the types
     in Quipp and their corresponding sufficient statistics, features, and English descriptions.
 
+\begin{table}[t]
+  \caption{Types in Quipp}
+  \label{type-table}
     \begin{center}
@@
 \end{center}
+\end{table}
 
@@
   Using this knowledge about the form of the random functions, we can improve the clustering model from before.
-  As observed in the previous graph, the two clusters are forced to have the
+  As observed in Figure 1, the two clusters are forced to have the
   same variance. They do not fit the data well, since the data has a different
@@
 \end{lstlisting}
-  Using this model, we get the following clusters:
+  \begin{figure}[h]
   \begin{center}
@@
   \end{center}
+  \caption{Clustering the Iris dataset with independent variances}
+\end{figure}
+  Using this model, we get the clusters shown in Figure 2.
   This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.
@@
   Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).
-This allows $x$ and $y$ to be correlated, improving fit:
+This allows $x$ and $y$ to be correlated, improving fit, as seen in Figure 3.
+  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.
 
+  \begin{figure}[h]
   \begin{center}
@@
   \end{center}
-
-  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.
+  \caption{Clustering the Iris dataset with multivariate Gaussian distributions}
+\end{figure}
 
@@
     Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
-    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.
+    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.  For the following examples,
+    for each iteration except the first, we plot quantiles of average regret per sample over multiple runs.  We exclude
+    the first iteration because it represents the regret of randomly generated parameters, which is much higher than the regret in future iterations.  Since
+    the expectation maximization algorithm does not always converge to a global optimum, we do not expect regret
+    to go to 0 as we increase the number of iterations.
 
@@
 
+  \begin{figure}[h]
 \begin{center}
@@
 \end{center}
+\caption{Regret-per-sample quantiles for clustering}
+\end{figure}
 
@@
 var model = function() {
-  var whichClass = randomValue(Class);
-  observe(classFeatures, whichClass);
+  var whichClass = observe(randomValue, Class);
+  observe(classFeatures[whichClass]);
 };
@@
 distribution for each.  Since each feature is boolean, we will learn
-a different categorical distribution for each class.
-
-(figure should show average classification accuracy)
+a different categorical distribution for each class.  The main difference from clustering is that,
+since this is a supervised learning method, we observe the class in addition to the features.  Since there
+are no latent variables, the system finds the correct parameters in a single iteration, so the regret plot is uninteresting.
 
@@
 \end{center}
+\caption{Regret-per-sample quantiles for factor analysis}
+\end{figure}
 
@@
 \end{center}
+\caption{Regret-per-sample quantiles for hidden Markov model}
+\end{figure}
 
@@
 \end{center}
+\caption{Regret-per-sample quantiles for latent Dirichlet allocation}
+\end{figure}
 \subsection{Neural network}
@@
 var model = function() {
-  var inputLayer = randomValue(Input);
+  var inputLayer = observe(randomValue, Input);
   var hiddenLayer = inputToHidden(inputs[sampIndex]);
@@
 Supporting disjoint union types is also straightforward.
-Also, it will be useful to create a more usable interface to infer parameters and perform additional data processing given these parameters.
+Also, it will be useful to create a more user-friendly interface for inferring parameters and performing additional data processing, such as inferring latent values given fixed parameters or visualizing the distributions implied by the parameters.
 

### 2015-06-05 plotaccuracy.py

@@ -5,8 +5,19 @@ import sys
 import json
 import subprocess
 
-
 program_name = sys.argv[1]
+
+nsamps = 1
+if program_name == 'nd_clustering':
+  nsamps = 100
+elif program_name == 'factor_analysis':
+  nsamps = 50
+elif program_name == 'hmm':
+  nsamps = 40
+elif program_name == 'lda':
+  nsamps = 40
+
+
 
 run_results = []
 for line in open('plots/' + program_name + '.data'):
@@ -38,6 +49,8 @@ def concat(lst):
 
 niters = len(run_results[0]['infLps'])
 
+niters = min(12, niters)
+
 def mean(xs):
   return sum(xs) / len(xs)
 
@@ -55,7 +68,7 @@ regret_distrs = [[r['origLp'] - r['infLps'][i] for r in run_results] for i in ra
 print 'rd', len(regret_distrs), niters
 
 xlabel('iteration')
-ylabel('regret')
+ylabel('regret per sample')
 
 quantiles = [0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9]
 quantiles = [0.1, 0.3, 0.5, 0.7, 0.9]
@@ -63,12 +76,12 @@ for q in quantiles:
   quants = []
   for i in range(burn, niters):
     regrets = regret_distrs[i-burn]
-    quants.append(max(0, quantile(regrets, q)))
-  plot(range(burn, niters), quants, label=str(q))
+    quants.append(max(0, quantile(regrets, q))/nsamps)
+  plot(range(burn+1, niters+1), quants, label=str(q))
   
 
-title('Regret quantiles by iteration')
 legend()
+axis(xmin=2)
 
 
 plot_name = 'plots/accuracy_' + program_name + '.png'

### 2015-06-05 plots/plotdata.py

@@ -80,13 +80,13 @@ def iparams3(px, py):
 
 
 
-which_plot = 2
+which_plot = 1
 
 if which_plot == 0:
 
-  xstdev0, xmean0, xmean1 = iparams1(params[0][0])
+  xstdev0, xmean0, xmean1 = iparams1(params1[0][0])
   xstdev1 = xstdev0
-  ystdev0, ymean0, ymean1 = iparams1(params[0][1])
+  ystdev0, ymean0, ymean1 = iparams1(params1[0][1])
   ystdev1 = ystdev0
 
 elif which_plot == 1:
@@ -101,20 +101,23 @@ else:
   xmean0, ymean0, angle0, ustdev0, vstdev0 = iparams3(params3[0], params3[2])
   xmean1, ymean1, angle1, ustdev1, vstdev1 = iparams3(params3[1], params3[3])
 
+fig = figure()
+ax = fig.add_subplot(111)
+ax.scatter(xs, ys)
 
 ellipses = []
 if which_plot != 2:
   for x,y,xs,ys,c in [(xmean0, ymean0,xstdev0, ystdev0, 'blue'), (xmean1, ymean1, xstdev1, ystdev1, 'red')]:
     ellipses.append((Ellipse(xy=(x,y), width=xs*2, height=ys*2), c))
 else:
-  for x,y,a,us,vs,c in [(xmean0, ymean0, angle0, ustdev0, vstdev0, 'blue'), (xmean1, ymean1, angle1, ustdev1, vstdev1, 'red')]:
+  for x,y,a,us,vs,c in [(xmean0, ymean0, angle0, ustdev0, vstdev0, 'red'), (xmean1, ymean1, angle1, ustdev1, vstdev1, 'blue')]:
     print 'xy', x, y, a, us, vs, c
     ellipses.append((Ellipse(xy=(x,y), width=us*2, height=vs*2, angle=a*180/math.pi), c))
 
-fig = figure()
-ax = fig.add_subplot(111)
 # ax.scatter(xs, ys, c=list(map(boolcolor, clusters)))
-ax.scatter(xs, ys)
+
+ylabel('Petal length (in)')
+xlabel('Sepal length (in)')
 
 for e,c in ellipses:
   ax.add_artist(e)

### 2015-06-05 src/expfam.js

@@ -65,7 +65,7 @@ var Double = {
     var mean = nps[0] * variance;
     return global.sample(s, k, a, erp.gaussianERP, [mean, Math.sqrt(variance)]);
   },
-  defaultNatParam: [0.0, -0.001],
+  defaultNatParam: [0.0, -0.5],
   randNatParam: fromMonad(function() {
     return mbind(global.sample, erp.gaussianERP, [0, 5], function(mean) {
       return mbind(global.sample, erp.gaussianERP, [0, 5], function(stdev) {
@@ -86,9 +86,6 @@ var Double = {
         });
       });
     });
-  }),
-  randomDefault: fromMonad(function() {
-    return mcurry(global.sample, erp.gaussianERP, [0, 1]);
   }),
   featuresMask: [true, false],
   mle: function(samples, natParam) {
@@ -132,13 +129,13 @@ function Categorical(n) {
     randParams: fromMonad(function(nfeatures) {
       var getRow = mcurry(randCategoricalNp, n);
       return mbind(getRow, function(base) {
+        if (nfeatures == 0) {
+          return mreturn({base: base, weights: _.times(n-1, function() { return []; })});
+        }
         return mbind(util.replicateM, nfeatures, getRow, function(weights) {
           return mreturn({base: base, weights: opt.transpose(weights)});
         });
       });
-    }),
-    randomDefault: fromMonad(function() {
-      return mcurry(global.sample, erp.randomIntegerERP, [n]);
     }),
     featuresMask: _.times(n-1, function() { return true; }),
     formatParams: function(params) { return params; }
@@ -203,9 +200,6 @@ function Tuple(types) {
         weights: util.concat(_.pluck(subParams, 'weights'))
       };
     },
-    randomDefault: mcurry(util.mapM, types, fromMonad(function(t) {
-      return t.randomDefault;
-    })),
     formatParams: function(params) {
       return types.map(function(type, i) {
         var subParams = {
@@ -311,7 +305,7 @@ function gaussianMle(samples) {
   ys.forEach(function(y, i) {
     var lp = logProbability(ad.standardNumType, Double, params, samples[i][1], samples[i][2]);
     var predLp = - 0.5 * Math.log(2 * Math.PI * resid2) - Math.pow(y - predYs[i], 2) / (2 * resid2); 
-    assert(Math.abs(lp - predLp) < 0.01);
[…]

### 2015-06-05 src/mh.js

@@ -168,6 +168,10 @@ module.exports = function(env) {
       }
       this.returnHist[stringifiedVal].prob += 1;
 
+      if (this.trace.length == 0) {
+        return this.exit(s, val);
+      }
+
       // make a new proposal:
       this.regenFrom = Math.floor(Math.random() * this.trace.length);
       var regen = this.trace[this.regenFrom];

### 2015-06-05 src/optimization.js

@@ -81,6 +81,7 @@ function elementsToMat(nrows, elems) {
 }
 
 function transpose(mat) {
+  assert (mat.length != 0);
   var res = [];
   for (var i = 0; i < mat[0].length; ++i) {
     res.push(matCol(mat, i));

### 2015-06-05 src/param_inference.js

@@ -32,11 +32,17 @@ var resetObsIndex = function(s, k, a) {
 DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var self = this;
   var args = [].slice.call(arguments, 2, arguments.length);
-  if (typeof addr == 'function' || typeof addr == 'object') {
+  if (typeof addr != 'string') {
     // return mbind(util.getAddress, function(a) {
     return mbind(nextObsIndex, function(a) {
-      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', '' + a, addr, rf].concat(args));
     });
+  }
+
+  if (typeof rf == 'number') {
+    assert(rf == Math.floor(rf));
+    self.valueMap[addr] = rf;
+    return mreturn(rf);
   }
 
   if (rf.constructor == erp.ERP) {
@@ -45,6 +51,7 @@ DataSampler.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
       return mreturn(samp);
     });
   }
+
 
   return mbind.apply(null, [rf].concat(args).concat([function(samp) {
     self.valueMap[addr] = samp;
@@ -70,18 +77,32 @@ function DataConditioner(valueMap) {
 DataConditioner.prototype.sampleOrCondition = fromMonad(function(addr, rf) {
   var self = this;
   var args = [].slice.call(arguments, 2, arguments.length);
-  if (typeof addr == 'function' || typeof addr == 'object') {
+  if (typeof addr != 'string') {
     return mbind(nextObsIndex, function(a) {
     // return mbind(util.getAddress, function(a) {
-      return mcurryMethod.apply(null, [self, 'sampleOrCondition', a, addr, rf].concat(args));
+      return mcurryMethod.apply(null, [self, 'sampleOrCondition', '' + a, addr, rf].concat(args));
     });
   }
 
   assert(addr in self.valueMap);
   var ret = self.valueMap[addr];
 
+  if (typeof rf == 'number') {
+    return mbind(factor, rf == ret ? 0.0 : -Infinity, function() {
+      return mreturn(rf);
+    });
+  }
+
   if (rf.constructor == erp.ERP) {
     return mbind(global.factor, rf.score(args, ret), function() {
+      return mreturn(ret);
+    });
+  }
+
+  if (rf == expfam.randomValue) {
+    var ef = args[0];
[…]

### 2015-06-05 src/quipp.js

@@ -31,5 +31,5 @@ module.exports = {
   Tuple: Tuple2,
   Vector: Vector,
   Categorical: util.makeWpplFunction(expfam.Categorical),
-  randomValue: fromMonad(function(t) { return t.randomDefault; })
+  randomValue: expfam.randomValue
 };

### 2015-06-05 src/util.js

@@ -53,9 +53,6 @@ function trampoline(f) {
   };
 }
 
-// 18008723377
-// 1800 traders
-
 function mbind(x) {
   assert(typeof x == 'function', x);
   var args = [].slice.call(arguments);

### 2015-06-05 paper/paper.tex
finishing up

@@
   applicability of our algorithm, by writing some of the most common machine learning models
-  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
-  a neural net.
+  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model and Latent Dirichlet Allocation.
 
@@
   In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
-  split the data into 2 clusters with equal variance, as seen in Figure 1.
+  split the data into 2 clusters with equal variance, as seen in Figure \ref{iris-orig}.
 
@@
   \caption{Clustering the Iris dataset}
+  \label{iris-orig}
 \end{figure}
@@
     Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
-    returns a sample from the ``default'' distribution for a given type.  Table 1 shows the types
+    returns a sample from the ``default'' distribution for a given type.  Table \ref{type-table} shows the types
     in Quipp and their corresponding sufficient statistics, features, and English descriptions.
@@
   Using this knowledge about the form of the random functions, we can improve the clustering model from before.
-  As observed in Figure 1, the two clusters are forced to have the
+  As observed in Figure \ref{iris-orig}, the two clusters are forced to have the
   same variance. They do not fit the data well, since the data has a different
@@
   \caption{Clustering the Iris dataset with independent variances}
+  \label{iris-indep}
 \end{figure}
-  Using this model, we get the clusters shown in Figure 2.
+Using this model, we get the clusters shown in Figure \ref{iris-indep}.
   This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.
@@
 Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).
-This allows $x$ and $y$ to be correlated, improving fit, as seen in Figure 3.
+This allows $x$ and $y$ to be correlated, improving fit, as seen in Figure \label{iris-dep}.
   This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.
@@
   \caption{Clustering the Iris dataset with multivariate Gaussian distributions}
+  \label{iris-dep}
 \end{figure}
@@
 
+\FloatBarrier
   \subsection{Clustering}
@@
 
-In this example, we cluster 2d points into 3 different clusters.  Given a cluster, the distribution for a point is some independent Gaussian distribution.  This is similar to fuzzy c-means clustering.
+In this example, we cluster 2d points into 3 different clusters.
+Given a cluster, the distribution for a point is some independent Gaussian distribution.
+This is similar to fuzzy c-means clustering.
+Regret usually reduces quite a lot as iterations increase.
 
@@
 
+\FloatBarrier
 \subsection{Naive Bayes}
@@
 
+\FloatBarrier
   \subsection{Factor analysis}
@@
 This results in the model attempting to predict each point as an affine function of a vector of standard normal values, with Gaussian noise.
+Regret reduces as iterations increase.
 
@@
 
+\FloatBarrier
   \subsection{Hidden Markov model}
@@
 This means that we will learn both the state transition matrix and the observation matrix.
+Regret starts low and does not change over time, indicating that only one EM iteration was necessary to accurately estimate the parameters.
 
@@
 
+\FloatBarrier
 \subsection{Latent Dirichlet allocation}
@@
 
-For latent Dirichlet allocation, we use the unknown function \texttt{topicToWord} to map latent topics to word distributions.  We will learn a different categorical distribution for each topic.
+For latent Dirichlet allocation, we use the unknown function \texttt{topicToWord} to map latent topics to word distributions.
+We will learn a different categorical distribution for each topic.
+Regret generally reduces over iterations.
 
@@
 \end{figure}
-\subsection{Neural network}
 
-{\small
-\begin{lstlisting}
-var Input = Vector(30, Bool);
-var Hidden = Vector(10, Bool);
-var Output = Bool;
-
-var inputToHidden = randFunction(Input, Hidden);
-var hiddenToOutput = randFunction(Hidden, Output);
-
-var model = function() {
-  var inputLayer = observe(randomValue, Input);
-  var hiddenLayer = inputToHidden(inputs[sampIndex]);
-  observe(hiddenToOutput, hiddenLayer);
-};
-\end{lstlisting}
-}
+% \FloatBarrier
+% \subsection{Neural network}
+% 
+% {\small
+% \begin{lstlisting}
+% var Input = Vector(30, Bool);
+% var Hidden = Vector(10, Bool);
+% var Output = Bool;
+% 
+% var inputToHidden = randFunction(Input, Hidden);
+% var hiddenToOutput = randFunction(Hidden, Output);
+% 
+% var model = function() {
+%   var inputLayer = observe(randomValue, Input);
+%   var hiddenLayer = inputToHidden(inputs[sampIndex]);
+%   observe(hiddenToOutput, hiddenLayer);
+% };
+% \end{lstlisting}
+% }
 

### 2015-06-05 plotaccuracy.py

@@ -49,7 +49,8 @@ def concat(lst):
 
 niters = len(run_results[0]['infLps'])
 
-niters = min(12, niters)
+if program_name == 'factor_analysis':
+  niters = min(12, niters)
 
 def mean(xs):
   return sum(xs) / len(xs)
